---
title: "03_EdgeR"
author: "Julin Maloof"
date: "2025-04-18"
output: html_document
---

Goal: TMM normalize all gametophyte samples together (NCBI and McConnell)
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(edgeR)
library(ggrepel)
```

## McConnel Counts

### Read the counts and create sample description
```{r}
counts.data.Mc <- read_csv("../../McConnell_Samples/output/ht-seq-counts-combined.csv.gz")
head(counts.data.Mc)
```

```{r}
counts.data.Mc %>% filter(str_detect(gene_ID, "__"))
```

```{r}
counts.data.Mc <- counts.data.Mc %>% filter(str_detect(gene_ID, "__", negate = TRUE))
```

```{r}
counts.data.Mc <- counts.data.Mc %>%
  rename_with(.fn = ~ str_remove(.x, "RESUB-")) %>%
  rename_with(.fn = ~ str_replace(.x, "(male|mature)-", "\\1_"))
```


```{r}
sample.description.Mc <- tibble(sample=colnames(counts.data.Mc)[-1]) %>%
  mutate(gt=str_extract(sample, "^[a-zA-Z0-9]*"),
         base_gt=str_remove(gt, "[0-9]+"),
         base_gt=str_replace(base_gt, "BA", "CrLFY1-OX"),
         base_gt=str_replace(base_gt, "B(D|F)", "CrLFY2-OX"),
         tissue=str_remove(sample, "^[a-zA-Z0-9]*-")) %>%
  separate(tissue, into=c("tissue", "rep"), sep = "-", fill="right", extra="merge") %>%
  mutate( group=str_c(base_gt, "-", tissue))

sample.description.Mc

write_csv(sample.description.Mc, "../input/McConnel.sample.description.Mc.csv")
```

### Filter for Gametophyte only

```{r}
sample.description.Mc <- sample.description.Mc %>% filter(str_detect(tissue, "gam|herm"))

counts.data.Mc <- counts.data.Mc %>% select(gene_ID, matches(sample.description.Mc$sample))
```

## NCBI counts

### Read the counts and create sample description
```{r}
counts.data.ncbi <- read_delim("../output/ht-seq-counts_ncbi.csv.gz")
head(counts.data.ncbi)
```

```{r}
counts.data.ncbi %>% filter(str_detect(gene_ID, "__"))
```

```{r}
counts.data.ncbi <- counts.data.ncbi %>% filter(str_detect(gene_ID, "__", negate = TRUE))
```


```{r}
sample.description.ncbi <- read_csv("../input/all_srr_info_short.csv")
sample.description.ncbi
```

what is missing?

```{r}
mapped.samples.ncbi <- tibble(Run=colnames(counts.data.ncbi)[-1])
sample.description.ncbi %>% anti_join(mapped.samples.ncbi)
```
callus.  not sure why this got dropped but probably okay.

```{r}
sample.description.ncbi <- sample.description.ncbi %>% semi_join(mapped.samples.ncbi)
```


### Filter for Gametophyte only

```{r}
sample.description.ncbi <- sample.description.ncbi %>% 
  filter(str_detect(stage, "gametophyte"),
         str_detect(description, "spore", negate = TRUE)) 

counts.data.ncbi <- counts.data.ncbi %>% select(gene_ID, matches(sample.description.ncbi$Run))
```

## Combine data sets


```{r}
sample.description <- sample.description.ncbi %>% 
  mutate(group=str_c(BioProject, description,sep = "-")) %>%
  select(sample=Run, group) %>% bind_rows(sample.description.Mc) %>%
  select(sample, group) %>%
  mutate(label=str_remove(group, "PRJ.*?-")) %>%
  group_by(label) %>%
  mutate(label.unique = str_c(label,"-",row_number()))

counts.data <- full_join(counts.data.ncbi, counts.data.Mc)

colnames(counts.data)[-1] == sample.description$sample

write_csv(sample.description, "../output/sample.description.gametophyte.all.csv")
```

remove low read genes:

```{r}
counts.data <- counts.data[rowSums(counts.data[,-1] > 5) >= 3,]
```

```{r}
counts.matrix <- counts.data %>% select(-gene_ID) %>% as.matrix()
rownames(counts.matrix) <- counts.data$gene_ID

dge.data <- DGEList(counts=counts.matrix, 
                    group=sample.description$group)
dim(dge.data) 
dge.data <- calcNormFactors(dge.data, method = "TMM")
dge.data$samples # look at the normalization factors
```


## MAKE AN MDS

```{r}
mdsvals <- plotMDS(dge.data, plot = FALSE) # get the MDS values for plotting
```


```{r}
mdsvals2 <- tibble(x=mdsvals$x, y=mdsvals$y, sample=rownames(dge.data$samples)) %>%
  inner_join(sample.description)


mdsvals2 %>% 
  ggplot(aes(x=x, y=y)) +
  theme_bw() +
  geom_point(size=3) +
  geom_text_repel(aes(label=label), size=2)
```

## Check LFY Levels

CrLFY1: Ceric.33G031700

CrLFY2: Ceric.18G076300

```{r}
CrLFY1 <- "Ceric.33G031700" 

CrLFY2 <- "Ceric.18G076300"
```


```{r}
cpm <- dge.data %>% cpm(log = FALSE) %>% as_tibble(rownames = "gene_ID")
```

### LFY1:

all samples
```{r}
cpm %>% filter(str_detect(gene_ID, CrLFY1)) %>%
  pivot_longer(-gene_ID, names_to = "sample", values_to = "cpm") %>%
  full_join(sample.description) %>%
  ggplot(aes(x=label.unique, y=cpm)) +
  geom_col() +
  theme(axis.text.x = element_text(angle=90, hjust = 1, vjust=.5)) +
  ggtitle("CrLFY1")
```
mean expression per group

```{r}
cpm %>% filter(str_detect(gene_ID, CrLFY1)) %>%
  pivot_longer(-gene_ID, names_to = "sample", values_to = "cpm") %>%
  full_join(sample.description) %>%
  group_by(group) %>%
  summarize(cpm=mean(cpm),
            sem=sd(cpm)/sqrt(n()),
            label=unique(label)) %>%
  ggplot(aes(x=label, y=cpm, ymax=cpm+sem, ymin=cpm-sem)) +
  geom_col() +
  geom_errorbar() +
  theme(axis.text.x = element_text(angle=90, hjust = 1, vjust=.5)) +
  ggtitle("CrLFY1")
```


### LFY2:

```{r}
cpm %>% filter(str_detect(gene_ID, CrLFY2)) %>%
  pivot_longer(-gene_ID, names_to = "sample", values_to = "cpm") %>%
  full_join(sample.description) %>%
  ggplot(aes(x=label.unique, y=cpm)) +
  geom_col() +
  theme(axis.text.x = element_text(angle=90, hjust = 1, vjust=.5)) +
  ggtitle("CrLFY2")
```
mean expression per group

```{r}
cpm %>% filter(str_detect(gene_ID, CrLFY2)) %>%
  pivot_longer(-gene_ID, names_to = "sample", values_to = "cpm") %>%
  full_join(sample.description) %>%
  group_by(group) %>%
  summarize(cpm=mean(cpm),
            sem=sd(cpm)/sqrt(n()),
            label=unique(label)) %>%
  ggplot(aes(x=label, y=cpm, ymax=cpm+sem, ymin=cpm-sem)) +
  geom_col() +
  geom_errorbar() +
  theme(axis.text.x = element_text(angle=90, hjust = 1, vjust=.5)) +
  ggtitle("CrLFY2")
```

## write out log counts

```{r}
dge.data %>% cpm(log = TRUE) %>% write.csv("../output/gam_combined_log2cpm.csv")
R.utils::gzip("../output/gam_combined_log2cpm.csv", overwrite=TRUE)
```

