---
title: "07_MADS_box"
author: "Julin Maloof"
date: "`r Sys.Date()`"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(topGO)
library(reticulate)
py_require("scipy")
library(fgsea)
conflicted::conflicts_prefer(dplyr::rename, dplyr::filter, dplyr::select)
```


Where are the MADS box genes relative to the LFY1 and LFY2 in the various networks that I have built?

```{r}
CrLFY1 <- "Ceric.33G031700"

CrLFY2 <- "Ceric.18G076300"
```


```{r}
MADS <- read_csv("../input/CrMADS_Carr.csv") # These are from Carr thesis
```

```{r}
MADS_new <- readxl::read_excel("../../shared/input/CrMADS_homologs_Crgenome-v2.1.xlsx") %>%
  mutate(geneID = str_remove(Locus, "\\.1$")) %>%
  select(geneID, everything())
```

```{r}
MADS$geneID %in% MADS_new$geneID
```

All of the previous MADS are in this list

Work with new list:

```{r}
MADS <- MADS_new %>% select(geneID, Result)
```


## WGCNA

### gametophyte samples
```{r}
load("../output/WGCNA_gametophyte_all_no4.5DAG.Rdata")
gmodule.assignment <- module.assignment %>%
  mutate(geneID = str_remove(geneID, "\\.v2.*"))
```

```{r}
table(gmodule.assignment$module)
```


```{r}
MADS.gmodules <- gmodule.assignment %>% right_join(MADS)
```

```{r}
table(MADS.gmodules$module)
```

```{r}
gmodule.assignment %>% filter(str_detect(geneID, CrLFY1))
```
```{r}
table(MADS.gmodules$module, MADS.gmodules$Result)
```

```{r}
gmodule.assignment %>% filter(str_detect(geneID, CrLFY2))
```

```{r}
MADS.gmodules %>% filter(str_detect(module, "black|brown"))
```

Test for sig. enrichment


```{r}
lfy1.gmodule <- gmodule.assignment %>% filter(str_detect(geneID, CrLFY1)) %>% pull(module)

lfy2.gmodule <- gmodule.assignment %>% filter(str_detect(geneID, CrLFY2)) %>% pull(module)

gfisherinput <- gmodule.assignment %>%
  mutate(lfy1.module = str_detect(module, lfy1.gmodule),
         lfy2.module = str_detect(module, lfy2.gmodule),
         mads.gene=geneID %in% MADS$geneID,
         mutate(across(-geneID, factor)))
```


```{r}
fisher.test(gfisherinput$lfy1.module, gfisherinput$mads.gene)
```
```{r}
fisher.test(gfisherinput$lfy2.module, gfisherinput$mads.gene)
```

### Sporophyte samples
```{r}
load("../output/WGCNA_sporophyte_all.Rdata")
smodule.assignment <- module.assignment %>%
  mutate(geneID = str_remove(geneID, "\\.v2.*"))
```

```{r}
table(smodule.assignment$module)
```


```{r}
MADS.smodules <- smodule.assignment %>% right_join(MADS)
```
```{r}
table(MADS.smodules$module)
```

```{r}
table(MADS.smodules$module, MADS.smodules$Result)
```

```{r}
smodule.assignment %>% filter(str_detect(geneID, CrLFY1))
```
```{r}
smodule.assignment %>% filter(str_detect(geneID, CrLFY2))
```
```{r}
MADS.smodules %>% filter(str_detect(module, "^blue$"))
```

```{r}
sfisherinput <- gmodule.assignment %>%
  mutate(lfy1.module = str_detect(module, lfy1.gmodule),
         lfy2.module = str_detect(module, lfy2.gmodule),
         mads.gene=geneID %in% MADS$geneID,
         mutate(across(-geneID, factor)))
```


```{r}
fisher.test(sfisherinput$lfy1.module, sfisherinput$mads.gene)
```
## TGCN Gametophyte

### LFY1 Cutoff 6

```{r}
read_csv("../output/TGCN_LFY1_gam_no4.5DAG/results/LFY1_gam_c6_TGCN.csv") %>% 
  mutate(geneID = str_remove(genes, "\\.[0-9]$")) %>%
  semi_join(MADS)
```

### LFY2 Cutoff 5

```{r}
read_csv("../output/TGCN_LFY2_gam/results/LFY2_gam_c5_TGCN.csv") %>%
  mutate(geneID = str_remove(genes, "\\.[0-9]$")) %>%
  semi_join(MADS)
```

## TGCN Sporophyte

### LFY1 Cutoff 6

```{r}
read_csv("../output/TGCN_LFY1_sp/results/LFY1_sp_c6_TGCN.csv") %>% 
  mutate(geneID = str_remove(genes, "\\.[0-9]$")) %>%
  semi_join(MADS)
```

### LFY2 Cutoff 5

```{r}
read_csv("../output/TGCN_LFY2_sp/results/LFY2_sp_c5_TGCN.csv") %>%
  mutate(geneID = str_remove(genes, "\\.[0-9]$")) %>%
  semi_join(MADS)
```

## Mutual Rank Networks

Create a mutual rank network and check for GSEA relative to MADS box

### Gametophyte
load sample info
```{r}
sample.description <- read.csv("../output/sample.description.gametophyte.no4.5DAG.csv")
```

load read counts

```{r}
lcpm <- read.csv("../output/gam_combined_log2cpm_no4.5DAG.csv.gz", row.names = 1, check.names = FALSE)

lcpm <- lcpm[,sample.description$sample]
rownames(lcpm) <- str_remove(rownames(lcpm), "\\.v2\\..")
head(lcpm)
dim(lcpm)
lcpmM <- as.matrix(lcpm)

lfy_idx <- c(which(rownames(lcpm) == CrLFY1), which(rownames(lcpm) == CrLFY2)) 

r_to_py(lcpmM)
r_to_py(lfy_idx)
```


```{python}
import numpy as np
from scipy.stats import rankdata

lfy_idx = np.array(r.lfy_idx, dtype=int) - 1

C = np.absolute(np.corrcoef(r.lcpmM, rowvar=True))        # shape (n_genes, n_genes)
np.fill_diagonal(C, 0.0)


# 2. rank correlations for each gene (higher corr = smaller rank)
#    here we rank by decreasing correlation, ignoring self-corr

n = C.shape[0]
ranks = np.zeros_like(C, dtype=float)


for i in range(n):
    # descending rank: biggest correlation -> rank 1
    rank = rankdata(-C[:, i], method='average')
    ranks[:, i] = rank
    
# 3. mutual rank: geometric mean of the two directional ranks
#    MR_ij = sqrt(rank_i(j) * rank_j(i))
MR = np.sqrt(ranks * ranks.T)

Clfy = C[:, lfy_idx]

MRlfy = MR[:, lfy_idx]

del C
del MR

```

```{r}
# Bring results back into R
cor.lfy.gam    <- py$Clfy
MR.lfy.gam   <- py$MRlfy

dimnames(cor.lfy.gam)  <- list(rownames(lcpm), c("CrLFY1", "CrLFY2"))
dimnames(MR.lfy.gam) <- list(rownames(lcpm), c("CrLFY1", "CrLFY2"))

cor.lfy.gam <- cor.lfy.gam %>% as_tibble(rownames = "geneID")
MR.lfy.gam <- MR.lfy.gam %>% as_tibble(rownames = "geneID")

head(cor.lfy.gam)
head(MR.lfy.gam)


```

#### GSEA

Create list for GSEA input
```{r}
MADS.list <- as.list(MADS) %>% set_names("MADS")
```

### MR

get ordered list for lfy1 and lfy2
```{r}
MR.lfy1.gam <-  MR.lfy.gam %>% arrange(CrLFY1) %>% pull(CrLFY1, name = geneID)
MR.lfy2.gam <- MR.lfy.gam %>% arrange(CrLFY2) %>% pull(CrLFY2, name = geneID)
```

```{r}
MR.lfy1.gsea.res.gam <- fgsea(MADS.list, MR.lfy1.gam, scoreType = "pos")
MR.lfy2.gsea.res.gam <- fgsea(MADS.list, MR.lfy2.gam, scoreType = "pos")
```

```{r}
MR.lfy1.gsea.res.gam
```

```{r}
MR.lfy2.gsea.res.gam
```

### correlation

get ordered list for lfy1 and lfy2
```{r}
cor.lfy1.gam <-  cor.lfy.gam %>% arrange(CrLFY1) %>% pull(CrLFY1, name = geneID)
cor.lfy2.gam <- cor.lfy.gam %>% arrange(CrLFY2) %>% pull(CrLFY2, name = geneID)
```

```{r}
cor.lfy1.gsea.res.gam <- fgsea(MADS.list, cor.lfy1.gam, scoreType = "pos")
cor.lfy2.gsea.res.gam <- fgsea(MADS.list, cor.lfy2.gam, scoreType = "pos")
```

```{r}
cor.lfy1.gsea.res.gam
```

```{r}
cor.lfy2.gsea.res.gam
```

### Sporophyte
load sample info
```{r}
sample.description <- read.csv("../output/sample.description.sporophyte.all.csv")
```

load read counts

```{r}
lcpm <- read.csv("../output/sporophyte_combined_log2cpm.csv.gz", row.names = 1, check.names = FALSE)

lcpm <- lcpm[,sample.description$sample]
rownames(lcpm) <- str_remove(rownames(lcpm), "\\.v2\\..")
head(lcpm)
dim(lcpm)
lcpmM <- as.matrix(lcpm)

lfy_idx <- c(which(rownames(lcpm) == CrLFY1), which(rownames(lcpm) == CrLFY2)) 

r_to_py(lcpmM)
r_to_py(lfy_idx)
```


```{python}
import numpy as np
from scipy.stats import rankdata

lfy_idx = np.array(r.lfy_idx, dtype=int) - 1

C = np.absolute(np.corrcoef(r.lcpmM, rowvar=True))        # shape (n_genes, n_genes)
np.fill_diagonal(C, 0.0)


# 2. rank correlations for each gene (higher corr = smaller rank)
#    here we rank by decreasing correlation, ignoring self-corr

n = C.shape[0]
ranks = np.zeros_like(C, dtype=float)


for i in range(n):
    # descending rank: biggest correlation -> rank 1
    rank = rankdata(-C[:, i], method='average')
    ranks[:, i] = rank
    
# 3. mutual rank: geometric mean of the two directional ranks
#    MR_ij = sqrt(rank_i(j) * rank_j(i))
MR = np.sqrt(ranks * ranks.T)

Clfy = C[:, lfy_idx]

MRlfy = MR[:, lfy_idx]

del C
del MR

```

```{r}
# Bring results back into R
cor.lfy.spor    <- py$Clfy
MR.lfy.spor   <- py$MRlfy

dimnames(cor.lfy.spor)  <- list(rownames(lcpm), c("CrLFY1", "CrLFY2"))
dimnames(MR.lfy.spor) <- list(rownames(lcpm), c("CrLFY1", "CrLFY2"))

cor.lfy.spor <- cor.lfy.spor %>% as_tibble(rownames = "geneID")
MR.lfy.spor <- MR.lfy.spor %>% as_tibble(rownames = "geneID")

head(cor.lfy.spor)
head(MR.lfy.spor)


```

#### GSEA


### MR

get ordered list for lfy1 and lfy2
```{r}
MR.lfy1.spor <-  MR.lfy.spor %>% arrange(CrLFY1) %>% pull(CrLFY1, name = geneID)
MR.lfy2.spor <- MR.lfy.spor %>% arrange(CrLFY2) %>% pull(CrLFY2, name = geneID)
```

```{r}
MR.lfy1.gsea.res.spor <- fgsea(MADS.list, MR.lfy1.spor, scoreType = "pos")
MR.lfy2.gsea.res.spor <- fgsea(MADS.list, MR.lfy2.spor, scoreType = "pos")
```

```{r}
MR.lfy1.gsea.res.spor
```

```{r}
MR.lfy2.gsea.res.spor
```

### correlation

get ordered list for lfy1 and lfy2
```{r}
cor.lfy1.spor <-  cor.lfy.spor %>% arrange(CrLFY1) %>% pull(CrLFY1, name = geneID)
cor.lfy2.spor <- cor.lfy.spor %>% arrange(CrLFY2) %>% pull(CrLFY2, name = geneID)
```

```{r}
cor.lfy1.gsea.res.spor <- fgsea(MADS.list, cor.lfy1.spor, scoreType = "pos")
cor.lfy2.gsea.res.spor <- fgsea(MADS.list, cor.lfy2.spor, scoreType = "pos")
```

```{r}
cor.lfy1.gsea.res.spor
```

```{r}
cor.lfy2.gsea.res.spor
```

# Mutual Rank Networks without ABS value

Create a mutual rank network and check for GSEA relative to MADS box

### Gametophyte
load sample info
```{r}
sample.description <- read.csv("../output/sample.description.gametophyte.no4.5DAG.csv")
```

load read counts

```{r}
lcpm <- read.csv("../output/gam_combined_log2cpm_no4.5DAG.csv.gz", row.names = 1, check.names = FALSE)

lcpm <- lcpm[,sample.description$sample]
rownames(lcpm) <- str_remove(rownames(lcpm), "\\.v2\\..")
head(lcpm)
dim(lcpm)
lcpmM <- as.matrix(lcpm)

lfy_idx <- c(which(rownames(lcpm) == CrLFY1), which(rownames(lcpm) == CrLFY2)) 

r_to_py(lcpmM)
r_to_py(lfy_idx)
```


```{python}
import numpy as np
from scipy.stats import rankdata

lfy_idx = np.array(r.lfy_idx, dtype=int) - 1

C = np.corrcoef(r.lcpmM, rowvar=True)        # shape (n_genes, n_genes)
np.fill_diagonal(C, 0.0)


# 2. rank correlations for each gene (higher corr = smaller rank)
#    here we rank by decreasing correlation, ignoring self-corr

n = C.shape[0]
ranks = np.zeros_like(C, dtype=float)


for i in range(n):
    # descending rank: biggest correlation -> rank 1
    rank = rankdata(-C[:, i], method='average')
    ranks[:, i] = rank
    
# 3. mutual rank: geometric mean of the two directional ranks
#    MR_ij = sqrt(rank_i(j) * rank_j(i))
MR = np.sqrt(ranks * ranks.T)

Clfy = C[:, lfy_idx]

MRlfy = MR[:, lfy_idx]

del C
del MR

```

```{r}
# Bring results back into R
cor.lfy.gam    <- py$Clfy
MR.lfy.gam   <- py$MRlfy

dimnames(cor.lfy.gam)  <- list(rownames(lcpm), c("CrLFY1", "CrLFY2"))
dimnames(MR.lfy.gam) <- list(rownames(lcpm), c("CrLFY1", "CrLFY2"))

cor.lfy.gam <- cor.lfy.gam %>% as_tibble(rownames = "geneID")
MR.lfy.gam <- MR.lfy.gam %>% as_tibble(rownames = "geneID")

head(cor.lfy.gam)
head(MR.lfy.gam)


```

#### GSEA

Create list for GSEA input
```{r}
MADS.list <- as.list(MADS) %>% set_names("MADS")
```

### MR

get ordered list for lfy1 and lfy2
```{r}
MR.lfy1.gam <-  MR.lfy.gam %>% arrange(CrLFY1) %>% pull(CrLFY1, name = geneID)
MR.lfy2.gam <- MR.lfy.gam %>% arrange(CrLFY2) %>% pull(CrLFY2, name = geneID)
```

```{r}
MR.lfy1.gsea.res.gam.pos <- fgsea(MADS.list, MR.lfy1.gam, scoreType = "pos")
MR.lfy2.gsea.res.gam.pos <- fgsea(MADS.list, MR.lfy2.gam, scoreType = "pos")
MR.lfy1.gsea.res.gam.neg <- fgsea(MADS.list, MR.lfy1.gam, scoreType = "neg")
MR.lfy2.gsea.res.gam.neg <- fgsea(MADS.list, MR.lfy2.gam, scoreType = "neg")
```

```{r}
MR.lfy1.gsea.res.gam.pos
```

```{r}
MR.lfy2.gsea.res.gam.pos
```
```{r}
MR.lfy1.gsea.res.gam.neg
```

```{r}
MR.lfy2.gsea.res.gam.neg
```
### correlation

get ordered list for lfy1 and lfy2
```{r}
cor.lfy1.gam <-  cor.lfy.gam %>% arrange(CrLFY1) %>% pull(CrLFY1, name = geneID)
cor.lfy2.gam <- cor.lfy.gam %>% arrange(CrLFY2) %>% pull(CrLFY2, name = geneID)
```

```{r}
cor.lfy1.gsea.res.gam.pos <- fgsea(MADS.list, cor.lfy1.gam, scoreType = "pos")
cor.lfy2.gsea.res.gam.pos <- fgsea(MADS.list, cor.lfy2.gam, scoreType = "pos")
cor.lfy1.gsea.res.gam.neg <- fgsea(MADS.list, cor.lfy1.gam, scoreType = "neg")
cor.lfy2.gsea.res.gam.neg <- fgsea(MADS.list, cor.lfy2.gam, scoreType = "neg")
```

```{r}
cor.lfy1.gsea.res.gam.pos
```

```{r}
cor.lfy2.gsea.res.gam.pos
```

```{r}
cor.lfy1.gsea.res.gam.neg
```

```{r}
cor.lfy2.gsea.res.gam.neg
```

### Sporophyte
load sample info
```{r}
sample.description <- read.csv("../output/sample.description.sporophyte.all.csv")
```

load read counts

```{r}
lcpm <- read.csv("../output/sporophyte_combined_log2cpm.csv.gz", row.names = 1, check.names = FALSE)

lcpm <- lcpm[,sample.description$sample]
rownames(lcpm) <- str_remove(rownames(lcpm), "\\.v2\\..")
head(lcpm)
dim(lcpm)
lcpmM <- as.matrix(lcpm)

lfy_idx <- c(which(rownames(lcpm) == CrLFY1), which(rownames(lcpm) == CrLFY2)) 

r_to_py(lcpmM)
r_to_py(lfy_idx)
```


```{python}
import numpy as np
from scipy.stats import rankdata

lfy_idx = np.array(r.lfy_idx, dtype=int) - 1

C = np.corrcoef(r.lcpmM, rowvar=True)        # shape (n_genes, n_genes)
np.fill_diagonal(C, 0.0)


# 2. rank correlations for each gene (higher corr = smaller rank)
#    here we rank by decreasing correlation, ignoring self-corr

n = C.shape[0]
ranks = np.zeros_like(C, dtype=float)


for i in range(n):
    # descending rank: biggest correlation -> rank 1
    rank = rankdata(-C[:, i], method='average')
    ranks[:, i] = rank
    
# 3. mutual rank: geometric mean of the two directional ranks
#    MR_ij = sqrt(rank_i(j) * rank_j(i))
MR = np.sqrt(ranks * ranks.T)

Clfy = C[:, lfy_idx]

MRlfy = MR[:, lfy_idx]

del C
del MR

```

```{r}
# Bring results back into R
cor.lfy.spor    <- py$Clfy
MR.lfy.spor   <- py$MRlfy

dimnames(cor.lfy.spor)  <- list(rownames(lcpm), c("CrLFY1", "CrLFY2"))
dimnames(MR.lfy.spor) <- list(rownames(lcpm), c("CrLFY1", "CrLFY2"))

cor.lfy.spor <- cor.lfy.spor %>% as_tibble(rownames = "geneID")
MR.lfy.spor <- MR.lfy.spor %>% as_tibble(rownames = "geneID")

head(cor.lfy.spor)
head(MR.lfy.spor)


```

#### GSEA


### MR

get ordered list for lfy1 and lfy2
```{r}
MR.lfy1.spor <-  MR.lfy.spor %>% arrange(CrLFY1) %>% pull(CrLFY1, name = geneID)
MR.lfy2.spor <- MR.lfy.spor %>% arrange(CrLFY2) %>% pull(CrLFY2, name = geneID)
```

```{r}
MR.lfy1.gsea.res.spor.pos <- fgsea(MADS.list, MR.lfy1.spor, scoreType = "pos")
MR.lfy2.gsea.res.spor.pos <- fgsea(MADS.list, MR.lfy2.spor, scoreType = "pos")
MR.lfy1.gsea.res.spor.neg <- fgsea(MADS.list, MR.lfy1.spor, scoreType = "neg")
MR.lfy2.gsea.res.spor.neg <- fgsea(MADS.list, MR.lfy2.spor, scoreType = "neg")
```

```{r}
MR.lfy1.gsea.res.spor.pos
```

```{r}
MR.lfy2.gsea.res.spor.pos
```

```{r}
MR.lfy1.gsea.res.spor.neg
```

```{r}
MR.lfy2.gsea.res.spor.neg
```

### correlation

get ordered list for lfy1 and lfy2
```{r}
cor.lfy1.spor <-  cor.lfy.spor %>% arrange(CrLFY1) %>% pull(CrLFY1, name = geneID)
cor.lfy2.spor <- cor.lfy.spor %>% arrange(CrLFY2) %>% pull(CrLFY2, name = geneID)
```

```{r}
cor.lfy1.gsea.res.spor.pos <- fgsea(MADS.list, cor.lfy1.spor, scoreType = "pos")
cor.lfy2.gsea.res.spor.pos <- fgsea(MADS.list, cor.lfy2.spor, scoreType = "pos")
cor.lfy1.gsea.res.spor.neg <- fgsea(MADS.list, cor.lfy1.spor, scoreType = "neg")
cor.lfy2.gsea.res.spor.neg <- fgsea(MADS.list, cor.lfy2.spor, scoreType = "neg")
```

```{r}
cor.lfy1.gsea.res.spor.pos
```

```{r}
cor.lfy2.gsea.res.spor.pos
```


```{r}
cor.lfy1.gsea.res.spor.neg
```

```{r}
cor.lfy2.gsea.res.spor.neg
```