---
title: "09_MR_intersection_with_gametophyte_sexDE"
output: html_notebook
---

Create a mutual rank network and intersect that with male vs hermaphrodite DE genes.  For this version, remove the male vs hermaphrodite samples from the MR/Correlation calculation.

```{r, message=FALSE}
library(tidyverse)
library(dbscan)    # Contains both DBSCAN and HDBSCAN implementations
library(patchwork)   # For arranging multiple plots
library(topGO)
library(broom)
conflicted::conflicts_prefer(dplyr::rename, dplyr::filter, dplyr::select)
```

```{r}
CrLFY1 <- "Ceric.33G031700.1" 

CrLFY2 <- "Ceric.18G076300.1"
```

load sample info
```{r}
sample.description <- read.csv("../output/sample.description.gametophyte.no4.5DAG.csv")
```

remove sex samples
```{r}
sample.description <- sample.description %>%
  filter(!str_detect(group, "WT-male_gam|WT-mature_herm"))
```

write the table back out for paper

```{r}
sample.description %>%
  mutate(rep = str_sub(label.unique, nchar(label.unique)),
         group = ifelse(str_detect(group, "^PRJ"), str_remove(group, "(-.*){1}$"), group)) %>%
  select(-label.unique) %>%
  rename(description=label) %>%
  rename_with(str_to_title) %>%
write_csv( "../output/sample.description.final.gametophyte.network.for.paper.csv")
```


load reads

```{r}
lcpm <- read.csv("../output/gam_combined_log2cpm_no4.5DAG.csv.gz", row.names = 1, check.names = FALSE)

lcpm <- lcpm[,sample.description$sample]
rownames(lcpm) <- str_remove(rownames(lcpm), fixed(".v2"))
head(lcpm)
dim(lcpm)
```

load DEG list for male vs hermaphrodite
```{r}
deg <- read_csv("../../McConnell_Samples/output/DEG_ALL_WT_male_v_herm.csv.gz") %>%
  slice_sample(n=nrow(.))
deg
```

Filter to keep the genes in the deg set
```{r}
lcpm <- lcpm[deg$geneID,]
```

Calculate MR matrix
```{r, eval=FALSE}
# correlation matrix
m.cor <- cor(t(lcpm))
diag(m.cor) <- 0

# rank matrix
mem.maxVSize(30000)
m.rank <- apply(m.cor,2,function(x) rank(-abs(x)))

# mutual rank matrix
MR <- sqrt(m.rank * t(m.rank))

# Create LFY specific subsets of these tables
cor.lfy <- m.cor[,c(CrLFY1, CrLFY2)] %>%
  round(3) %>%
  as_tibble(rownames = "geneID") %>%
  rename(LFY1.cor = contains(CrLFY1), LFY2.cor = contains(CrLFY2))

MR.lfy <- MR[,c(CrLFY1, CrLFY2)] %>%
  as_tibble(rownames = "geneID") %>%
  rename(LFY1.MR = contains(CrLFY1), LFY2.MR = contains(CrLFY2))

# Combine DEGs with LFY correlations and MR
deg.cor <- deg %>% right_join(cor.lfy) %>% 
  left_join(MR.lfy) %>%
  mutate(LFY1.rank = rank(-abs(LFY1.cor)),
         LFY2.rank = rank(-abs(LFY2.cor))) %>%
  select(geneID:FDR, starts_with("LFY"), everything()) %>%
  mutate(FDR.lt.0.05=FDR<0.05)

write_csv(deg.cor, "../../McConnell_Samples/output/DEG_ALL_WT_male_v_herm_LFY_CORRELATIONS_v2_noSexMR.csv.gz")
```

```{r}
deg.cor <- read_csv("../../McConnell_Samples/output/DEG_ALL_WT_male_v_herm_LFY_CORRELATIONS_v2_noSexMR.csv.gz")
```


```{r}
deg.cor %>%
  ggplot(aes(x=logFC, y=LFY1.cor)) +
  geom_point(alpha=.05)

ggsave("../output/LFY1_cor_logFC.png", width=6, height=4)
```

```{r}
deg.cor %>%
  ggplot(aes(x=logFC, y=LFY2.cor)) +
  geom_point(alpha=.1)
```

```{r}
deg.cor %>%
  ggplot(aes(x=LFY1.MR, y=logFC)) +
  geom_point(alpha=.1)
```

```{r}
deg.cor %>%
  ggplot(aes(x=LFY1.rank, y=logFC)) +
  geom_point(alpha=.1)
```

```{r}
deg.cor %>%
  ggplot(aes(x=LFY2.MR, y=logFC)) +
  geom_point(alpha=.1)
```


```{r}
deg.cor %>%
  ggplot(aes(x=LFY2.rank, y=logFC)) +
  geom_point(alpha=.1)
```

```{r}
deg.cor %>%
  arrange(LFY1.rank) %>%
  slice_tail(n=14000)
```

## Color points by WGCNA Cluster

```{r}
# Load WGCNA module assignment
load("../output/WGCNA_gametophyte_all_no4.5DAG_noSex.Rdata")

module.assignment <- module.assignment %>%
  mutate(geneID = str_remove(geneID, fixed(".v2"))) %>%
  select(geneID, module)
```

```{r}
pl.df <- deg.cor %>%
  left_join(module.assignment, by = c("geneID" = "geneID")) %>%
  mutate(module = ifelse(is.na(module), "grey80", module)) %>%
  select(geneID, module, everything())

pl.df %>%
  ggplot(aes(x = LFY1.MR, y = logFC)) +
  geom_point(color = pl.df$module, alpha = 0.1) 
```

```{r}
pl.df <- deg.cor %>%
  left_join(module.assignment, by = c("geneID" = "geneID")) %>%
  mutate(module = ifelse(is.na(module), "none", module),
         module = ifelse(module=="turquoise", "turquoise", "grey95")) %>%
  select(geneID, module, everything()) %>%
  arrange((module))

pl.df %>%
  ggplot(aes(x = LFY1.MR, y = logFC)) +
  geom_point(color = pl.df$module, alpha = 0.3) + theme_minimal()
```

## Try to ID clusters

RANK
```{r}
# 1. DBSCAN approach

# Normalize/scale data (optional but recommended)
# This helps when variables have different ranges
scaled_data <- scale(deg.cor[, c("logFC", "LFY1.rank")])

# Optionally, find the optimal DBSCAN parameters using kNNdistplot
# This helps determine a good eps value.  K should be minpt -1 
kNNdistplot(scaled_data[, c("logFC", "LFY1.rank")], k = 499)
abline(h = 0.2, col = "red") # Adjust this line to find the "elbow"

dbscan_result <- dbscan(scaled_data[, c("logFC", "LFY1.rank")], eps = 0.25, minPts = 500, borderPoints = TRUE)
deg.cor$dbscan_cluster <- generics::as.factor(dbscan_result$cluster)

# Create plots to compare the results
plot1 <- ggplot(deg.cor, aes(x = LFY1.rank, y = logFC,  color = dbscan_cluster)) +
  geom_point(alpha = 0.1) +
  #scale_color_brewer(name = "Cl", type = "qual") +
  ggtitle(str_c("DBSCAN Clustering, border points, eps: 0.25")) +
  theme_minimal()

print(plot1)

hdbscan_result <- hdbscan(scaled_data[, c("logFC", "LFY1.rank")], minPts = 500)
deg.cor$hdbscan_cluster <- generics::as.factor(hdbscan_result$cluster)

plot2 <- ggplot(deg.cor, aes(x = LFY1.rank, y = logFC, color = hdbscan_cluster)) +
  geom_point(alpha = 0.1) +
  scale_color_discrete(name = "HDBSCAN Cluster") +
  ggtitle("HDBSCAN Clustering") +
  theme_minimal()

print(plot2)

```

MR
```{r}
# 1. DBSCAN approach

# Normalize/scale data (optional but recommended)
# This helps when variables have different ranges
scaled_data <- scale(deg.cor[, c("logFC", "LFY1.MR")])

# Optionally, find the optimal DBSCAN parameters using kNNdistplot
# This helps determine a good eps value.  K should be minpt -1 
kNNdistplot(scaled_data[, c("logFC", "LFY1.MR")], k = 499)
abline(h = 0.2, col = "red") # Adjust this line to find the "elbow"

dbscan_result <- dbscan(scaled_data[, c("logFC", "LFY1.MR")], eps = 0.25, minPts = 500, borderPoints = TRUE)
deg.cor$dbscan_cluster <- generics::as.factor(dbscan_result$cluster)

# Create plots to compare the results
plot1 <- deg.cor %>% 
  arrange(dbscan_cluster) %>%
  ggplot(aes(x = LFY1.MR, y = logFC, color = dbscan_cluster)) +
  geom_point(alpha = 0.2) +
  scale_color_brewer(name = "Cl", type = "qual") +
  ggtitle(str_c("DBSCAN Clustering, border points, eps: 0.25")) +
  guides(color = guide_legend(override.aes = list(alpha = 1)))


print(plot1)

plot1.alt <- deg.cor %>% 
  arrange(desc(dbscan_cluster)) %>%
  ggplot(aes(x = LFY1.MR, y = logFC, color = dbscan_cluster)) +
  geom_point(alpha = 0.2) +
  scale_color_brewer(name = "Cl", type = "qual") +
  ggtitle(str_c("DBSCAN Clustering, border points, eps: 0.25")) +
  guides(color = guide_legend(override.aes = list(alpha = 1)))


print(plot1.alt)

hdbscan_result <- hdbscan(scaled_data[, c("logFC", "LFY1.MR")], minPts = 500)
deg.cor$hdbscan_cluster <- generics::as.factor(hdbscan_result$cluster)

plot2 <- deg.cor %>% ggplot(aes(x = LFY1.MR, y = logFC, color = hdbscan_cluster)) +
  geom_point(alpha = 0.2) +
  #scale_color_discrete(name = "HDBSCAN Cluster") +
  ggtitle("HDBSCAN Clustering") +
  scale_color_brewer(name = "Cl", type = "qual") +
  guides(color = guide_legend(override.aes = list(alpha = 1)))


print(plot2)

```
Mutual Rank DBSCAN is the best...

Create a column for being in cluster 2 and save output
```{r}
deg.cor <- deg.cor %>%
  mutate(in_LFY1_cluster = dbscan_cluster == 2)

deg.cor %>%
  select(geneID, in_LFY1_cluster, everything(), -starts_with("LFY2")) %>%
  write_csv("../output/LFY1_gam_sex_cluster_noSexSamples.csv.gz")
```


```{r}
deg.cor %>%
  arrange(desc(in_LFY1_cluster), logFC) %>%
  select(geneID, logFC, FDR, LFY1.cor, LFY1.MR, LFY1.rank, in_LFY1_cluster)
```

Where is this cluster on the cor.plot?

```{r}
deg.cor %>%
  ggplot(aes(x = logFC, y = LFY1.cor, color = dbscan_cluster)) +
  geom_point(alpha = 0.1) +
  scale_color_brewer(name = "Cl", type = "qual") +
  guides(color = guide_legend(override.aes = list(alpha = 1)))
```


## Compare LFY1 and LFY2 MR in cluster

```{r}
deg.cor %>%
  mutate(upMale = logFC < -1 & FDR < 0.01) %>%
  t.test(LFY1.MR ~ upMale, data = .)

deg.cor %>%
  mutate(upMale = logFC < -1 & FDR < 0.01) %>%
  t.test(LFY2.MR ~ upMale, data = .)
```
```{r}
deg.cor %>%
  t.test(LFY1.MR ~ in_LFY1_cluster, data = .)

deg.cor %>%
  t.test(LFY2.MR ~ in_LFY1_cluster, data = .)
```

```{r}
lfy.network.threshold <- deg.cor %>% filter(in_LFY1_cluster) %>% pull(LFY1.MR) %>% max()

fisher.tmp <- deg.cor %>% mutate(LFY1.network = LFY1.MR <= lfy.network.threshold,
                   LFY2.network = LFY2.MR <= lfy.network.threshold,
                   upMale = logFC < 0 & FDR < 0.01) 

fisher.test(fisher.tmp$upMale, fisher.tmp$LFY1.network, alternative = "greater")
fisher.test(fisher.tmp$upMale, fisher.tmp$LFY2.network, alternative = "greater")

```


## GO enrichment for LFY1 cluster

```{r}
gene2GO.list <- deg.cor %>% 
  rowwise() %>%
  mutate(GOlist = str_split(GO, pattern = " ")) %>%
  ungroup() %>%
  mutate(GOlist = set_names(GOlist, geneID)) %>%
  pull(GOlist)

cluster.genes <- deg.cor %>%
  mutate(cluster.genes = as.numeric(in_LFY1_cluster)) %>% 
  pull(cluster.genes, name = geneID) %>% generics::as.factor()

go.data.cluster <- new("topGOdata", ontology = "BP", description = "In LFY1 Cluster", allGenes = cluster.genes, annot = annFUN.gene2GO, gene2GO = gene2GO.list)

result.cluster <- runTest(go.data.cluster, algorithm = "weight01", statistic = "fisher")

GO.results <- GenTable(go.data.cluster, result.cluster, topNodes = sum(result.cluster@score < 0.05 )) %>%
  dplyr::rename_with( \(x) str_replace(x, "result1|apply.*", "p.value")) %>%
  mutate(p.value = as.numeric(p.value))

GO.results
```

```{r, message=FALSE}
write_csv(GO.results, file = "../output/LFY1_Sex_Cluster_GO_noSexSamples.csv")
```

## Look for overlap with A.t. repro  genes in all male vs hermaphrodite DEG

load gene info

```{r}
Cr_gene_info <- read_csv("../../shared/output/Cr_At_repro_Placket.csv")
```


### Enrichment

For each gene set (up or down), retain genes that are orthologs of A.t. vegetative or one of the reproductive sets and then do Fisher's exact.

```{r}
Cr_gene_info_deg <- deg %>% 
  select(geneID, logFC, FDR) %>%
  inner_join(Cr_gene_info, by=c("geneID"="Crichardii")) %>%
  select(-Athaliana) %>%
  mutate(upHermaphrodite = FDR < 0.01 & logFC > 0,
         upMale = FDR < 0.01 & logFC < 0) %>%
  select(geneID, starts_with("up"), everything(), -FDR, -logFC)

Cr_gene_info_deg


```

Is it ratio of each reproductive category to all others, or to vegetative specific?


```{r}
get.fisher <- function(category, module, t = Cr_gene_info_deg) {
  fisher.test(get(category, t), get(module, t), alternative = "greater")
}

fisher.results <- expand_grid(gene.category = colnames(Cr_gene_info)[-1:-2],
                              module = c("upMale", "upHermaphrodite")) %>%
  mutate(fisher = map2(gene.category, module, get.fisher, Cr_gene_info_deg),
         fisher.tidy = map(fisher, tidy)) %>%
  unnest(fisher.tidy) %>%
  rename(fisher.p.value = p.value) %>%
    mutate(fisher.fdr = p.adjust(fisher.p.value, method = "fdr")) %>%
  arrange(fisher.p.value)

```

```{r}
fisher.results.filter <- fisher.results %>%
  # group_by(module) %>%
  ungroup() %>%
  filter(fisher.fdr < 0.05) %>%
  arrange(module, fisher.fdr) %>%
  select(module, gene.category, estimate, fisher.fdr, everything(), -fisher, -method, -alternative)

#Table 1
write_csv(fisher.results.filter, "../output/male_hermaphrodite_DEG_At_repro_fisher_fdr_0.05.csv")

# Supplemental Table 1
write_csv(fisher.results, "../output/male_hermaphrodite_DEG_At_repro_fisher_all.csv")


fisher.results.filter 
```


## Look for overlap with A.t. repro  genes in male vs hermaphrodite DEG LFY1 cluster

### Enrichment

For each gene set (up or down), retain genes that are orthologs of A.t. vegetative or one of the reproductive sets and then do Fisher's exact.

```{r}
Cr_gene_info_LFY1Cluster <- deg.cor %>% 
  select(geneID, in_LFY1_cluster) %>%
  inner_join(Cr_gene_info, by=c("geneID"="Crichardii")) %>%
  select(-Athaliana)

Cr_gene_info_LFY1Cluster


```

Is it ratio of each reproductive category to all others, or to vegetative specific?


```{r}
get.fisher <- function(category, module, t = Cr_gene_info_LFY1Cluster) {
  fisher.test(get(category, t), get(module, t), alternative = "greater")
}

fisher.results.LFY1.cluster <- expand_grid(gene.category = colnames(Cr_gene_info)[-1:-2],
                              module = "in_LFY1_cluster") %>%
  mutate(fisher = map2(gene.category, module, get.fisher, Cr_gene_info_LFY1Cluster),
         fisher.tidy = map(fisher, tidy)) %>%
  unnest(fisher.tidy) %>%
  rename(fisher.p.value = p.value) %>%
    mutate(fisher.fdr = p.adjust(fisher.p.value, method = "fdr")) %>%
  arrange(fisher.p.value)

```

```{r}
fisher.results.LFY1.cluster.filter <- fisher.results.LFY1.cluster %>%
  # group_by(module) %>%
  ungroup() %>%
  filter(fisher.fdr < 0.05) %>%
  arrange(module, fisher.fdr) %>%
  select(module, gene.category, estimate, fisher.fdr, everything(), -fisher, -method, -alternative)

write_csv(fisher.results.LFY1.cluster.filter, "../output/male_hermaphrodite_DEG_LFY1_Cluster_At_repro_fisher.csv")

fisher.results.LFY1.cluster.filter
```