---
title: "09.3_Sporophyte_MR_intersection_with_gametophyte_sexDE"
output: html_notebook
---

Create a mutual rank network and intersect that with male vs hermaphrodite DE genes. Here we see if the Sporophyte MR network has any interesting patterns with (gametophyte) sex DE

```{r, message=FALSE}
library(tidyverse)
library(dbscan)    # Contains both DBSCAN and HDBSCAN implementations
library(patchwork)   # For arranging multiple plots
library(topGO)
conflicted::conflicts_prefer(dplyr::rename, dplyr::filter, dplyr::select)
```

```{r}
CrLFY1 <- "Ceric.33G031700.1" 

CrLFY2 <- "Ceric.18G076300.1"
```

load sample info
```{r}
sample.description <- read.csv("../output/sample.description.sporophyte.all.csv")
```

load reads

```{r}
lcpm <- read.csv("../output/sporophyte_combined_log2cpm.csv.gz", row.names = 1, check.names = FALSE)

lcpm <- lcpm[,sample.description$sample]
rownames(lcpm) <- str_remove(rownames(lcpm), fixed(".v2"))
head(lcpm)
dim(lcpm)
```

load DEG list for male vs hermaphrodite
```{r}
deg <- read_csv("../../McConnell_Samples/output/DEG_ALL_WT_male_v_herm.csv.gz") %>%
  slice_sample(n=nrow(.))
deg
```

Filter to keep the genes in the deg set
```{r}
lcpm <- lcpm[deg$geneID,]
```

correlation matrix
```{r}
m.cor <- cor(t(lcpm))
diag(m.cor) <- 0
```

rank matrix
```{r}
mem.maxVSize(30000)
m.rank <- apply(m.cor,2,function(x) rank(-abs(x)))
```

MR matrix
```{r}
MR <- sqrt(m.rank * t(m.rank))
MR[1:10,1:10]
```
Create LFY specific subsets of these tables
```{r}
cor.lfy <- m.cor[,c(CrLFY1, CrLFY2)] %>%
  round(3) %>%
  as_tibble(rownames = "geneID") %>%
  rename(LFY1.cor = contains(CrLFY1), LFY2.cor = contains(CrLFY2))

MR.lfy <- MR[,c(CrLFY1, CrLFY2)] %>%
  as_tibble(rownames = "geneID") %>%
  rename(LFY1.MR = contains(CrLFY1), LFY2.MR = contains(CrLFY2))
```

combine them
```{r}
deg.cor <- deg %>% right_join(cor.lfy) %>% 
  left_join(MR.lfy) %>%
  mutate(LFY1.rank = rank(-abs(LFY1.cor)),
         LFY2.rank = rank(-abs(LFY2.cor))) %>%
  select(geneID:FDR, starts_with("LFY"), everything()) %>%
  mutate(FDR.lt.0.05=FDR<0.05)
```

```{r}
deg.cor
```


```{r}
deg.cor %>%
  ggplot(aes(x=logFC, y=LFY1.cor)) +
  geom_point(alpha=.05)

```

```{r}
deg.cor %>%
  ggplot(aes(x=logFC, y=LFY2.cor)) +
  geom_point(alpha=.1)
```

```{r}
deg.cor %>%
  ggplot(aes(x=LFY1.MR, y=logFC)) +
  geom_point(alpha=.1)
```

```{r}
deg.cor %>%
  ggplot(aes(x=LFY1.rank, y=logFC)) +
  geom_point(alpha=.1)
```

```{r}
deg.cor %>%
  ggplot(aes(x=LFY2.MR, y=logFC)) +
  geom_point(alpha=.1)
```


```{r}
deg.cor %>%
  ggplot(aes(x=LFY2.rank, y=logFC)) +
  geom_point(alpha=.1)
```

```{r}
deg.cor %>%
  arrange(LFY1.rank) %>%
  slice_tail(n=14000)
```

## Color points by WGCNA Cluster

```{r}
# Load WGCNA module assignment
load("../output/WGCNA_sporophyte_all.Rdata")

module.assignment <- module.assignment %>%
  mutate(geneID = str_remove(geneID, fixed(".v2"))) %>%
  select(geneID, module)
```

```{r}
pl.df <- deg.cor %>%
  left_join(module.assignment, by = c("geneID" = "geneID")) %>%
  mutate(module = ifelse(is.na(module), "grey80", module)) %>%
  select(geneID, module, everything())

pl.df %>%
  ggplot(aes(x = LFY1.MR, y = logFC)) +
  geom_point(color = pl.df$module, alpha = 0.1) 
```

```{r}
pl.df <- deg.cor %>%
  left_join(module.assignment, by = c("geneID" = "geneID")) %>%
  mutate(module = ifelse(is.na(module), "none", module),
         module = ifelse(module=="turquoise", "blue", "red")) %>%
  select(geneID, module, everything()) %>%
  arrange(desc(module))

pl.df %>%
  ggplot(aes(x = LFY1.MR, y = logFC)) +
  geom_point(color = pl.df$module, alpha = 0.1) 
```

## Try to ID clusters

RANK
```{r}
# 1. DBSCAN approach

# Normalize/scale data (optional but recommended)
# This helps when variables have different ranges
scaled_data <- scale(deg.cor[, c("logFC", "LFY1.rank")])

# Optionally, find the optimal DBSCAN parameters using kNNdistplot
# This helps determine a good eps value.  K should be minpt -1 
kNNdistplot(scaled_data[, c("logFC", "LFY1.rank")], k = 499)
abline(h = 0.2, col = "red") # Adjust this line to find the "elbow"

dbscan_result <- dbscan(scaled_data[, c("logFC", "LFY1.rank")], eps = 0.25, minPts = 500, borderPoints = TRUE)
deg.cor$dbscan_cluster <- as.factor(dbscan_result$cluster)

# Create plots to compare the results
plot1 <- ggplot(deg.cor, aes(x = LFY1.rank, y = logFC,  color = dbscan_cluster)) +
  geom_point(alpha = 0.1) +
  #scale_color_brewer(name = "Cl", type = "qual") +
  ggtitle(str_c("DBSCAN Clustering, border points, eps: 0.25")) +
  theme_minimal()

print(plot1)

hdbscan_result <- hdbscan(scaled_data[, c("logFC", "LFY1.rank")], minPts = 500)
deg.cor$hdbscan_cluster <- as.factor(hdbscan_result$cluster)

plot2 <- ggplot(deg.cor, aes(x = LFY1.rank, y = logFC, color = hdbscan_cluster)) +
  geom_point(alpha = 0.1) +
  scale_color_discrete(name = "HDBSCAN Cluster") +
  ggtitle("HDBSCAN Clustering") +
  theme_minimal()

print(plot2)

```

MR
```{r}
# 1. DBSCAN approach

# Normalize/scale data (optional but recommended)
# This helps when variables have different ranges
scaled_data <- scale(deg.cor[, c("logFC", "LFY1.MR")])

# Optionally, find the optimal DBSCAN parameters using kNNdistplot
# This helps determine a good eps value.  K should be minpt -1 
kNNdistplot(scaled_data[, c("logFC", "LFY1.MR")], k = 499)
abline(h = 0.2, col = "red") # Adjust this line to find the "elbow"

dbscan_result <- dbscan(scaled_data[, c("logFC", "LFY1.MR")], eps = 0.25, minPts = 500, borderPoints = TRUE)
deg.cor$dbscan_cluster <- as.factor(dbscan_result$cluster)

# Create plots to compare the results
plot1 <- ggplot(deg.cor, aes(x = LFY1.MR, y = logFC, color = dbscan_cluster)) +
  geom_point(alpha = 0.2) +
  scale_color_brewer(name = "Cl", type = "qual") +
  ggtitle(str_c("DBSCAN Clustering, border points, eps: 0.25")) +
  guides(color = guide_legend(override.aes = list(alpha = 1)))


print(plot1)

hdbscan_result <- hdbscan(scaled_data[, c("logFC", "LFY1.MR")], minPts = 500)
deg.cor$hdbscan_cluster <- as.factor(hdbscan_result$cluster)

plot2 <- ggplot(deg.cor, aes(x = LFY1.MR, y = logFC, color = hdbscan_cluster)) +
  geom_point(alpha = 0.2) +
  #scale_color_discrete(name = "HDBSCAN Cluster") +
  ggtitle("HDBSCAN Clustering") +
  scale_color_brewer(name = "Cl", type = "qual") +
  guides(color = guide_legend(override.aes = list(alpha = 1)))


print(plot2)

```
Mutual Rank DBSCAN is the best...

Create a column for being in cluster 2 and save output
```{r}
deg.cor <- deg.cor %>%
  mutate(in_LFY1_cluster = dbscan_cluster == 2) %>%
  select(geneID, in_LFY1_cluster, everything(), -starts_with("LFY2"))

write_csv(deg.cor, "../output/LFY1_spor_gam_sex_cluster.csv")
```


```{r}
deg.cor %>%
  arrange(desc(in_LFY1_cluster), logFC) %>%
  select(geneID, logFC, FDR, LFY1.cor, LFY1.MR, LFY1.rank, in_LFY1_cluster)
```

## GO enrichment

```{r}
gene2GO.list <- deg.cor %>% 
  rowwise() %>%
  mutate(GOlist = str_split(GO, pattern = " ")) %>%
  ungroup() %>%
  mutate(GOlist = set_names(GOlist, geneID)) %>%
  pull(GOlist)

cluster.genes <- deg.cor %>%
  mutate(cluster.genes = as.numeric(in_LFY1_cluster)) %>% 
  pull(cluster.genes, name = geneID) %>% as.factor()

go.data.cluster <- new("topGOdata", ontology = "BP", description = "In LFY1 Cluster", allGenes = cluster.genes, annot = annFUN.gene2GO, gene2GO = gene2GO.list)

result.cluster <- runTest(go.data.cluster, algorithm = "weight01", statistic = "fisher")

GO.results <- GenTable(go.data.cluster, result.cluster, topNodes = sum(result.cluster@score < 0.05 )) %>%
  dplyr::rename_with( \(x) str_replace(x, "result1|apply.*", "p.value")) %>%
  mutate(p.value = as.numeric(p.value))

GO.results
```

```{r, message=FALSE}
write_csv(GO.results, file = "../output/LFY1_spor_Sex_Cluster_GO.csv")
```

