---
title: "09.4_MR_intersection_with_gametophyte_Atallah_DE"
output: html_notebook
---

Create a mutual rank network and intersect that with Atallah's antheridogen vs hermaphrodite DE genes.  

```{r, message=FALSE}
library(tidyverse)
library(dbscan)    # Contains both DBSCAN and HDBSCAN implementations
library(patchwork)   # For arranging multiple plots
library(topGO)
library(broom)
conflicted::conflicts_prefer(dplyr::rename, dplyr::filter, dplyr::select)
```

```{r}
CrLFY1 <- "Ceric.33G031700.1" 

CrLFY2 <- "Ceric.18G076300.1"
```

load sample info
```{r}
sample.description <- read.csv("../output/sample.description.gametophyte.no4.5DAG.csv")
```

load DEG list for Atallah
```{r}
deg <- read_csv("../output/DEG_ALL_Atallah.csv.gz") %>%
  slice_sample(n=nrow(.))
deg
```

load MR network.  See script 09.2 for creation.  This currently has McConnell DEG info in it and that will need to be replaced
```{r}
deg.cor <- 
  read_csv("../../McConnell_Samples/output/DEG_ALL_WT_male_v_herm_LFY_CORRELATIONS_v2_noSexMR.csv.gz") %>%
  select(-(logFC:FDR)) %>% # remove McConnell data
  
# Combine Atallah DEGs with LFY correlations and MR
  left_join(deg) %>%
  select(geneID:FDR, starts_with("LFY"), everything())

deg.cor
```


```{r}
deg.cor %>%
  ggplot(aes(x=logFC, y=LFY1.cor)) +
  geom_point(alpha=.05)
```

```{r}
deg.cor %>%
  ggplot(aes(x=logFC, y=LFY2.cor)) +
  geom_point(alpha=.1)
```

```{r}
deg.cor %>%
  ggplot(aes(x=LFY1.MR, y=logFC)) +
  geom_point(alpha=.1)
```

```{r}
deg.cor %>%
  ggplot(aes(x=LFY1.rank, y=logFC)) +
  geom_point(alpha=.1)
```

```{r}
deg.cor %>%
  ggplot(aes(x=LFY2.MR, y=logFC)) +
  geom_point(alpha=.1)
```


```{r}
deg.cor %>%
  ggplot(aes(x=LFY2.rank, y=logFC)) +
  geom_point(alpha=.1)
```

## color points by McConnell LFY1 Cluster

Create a column for being in cluster 2 and save output
```{r}
mcconnell_lfy_cluster <- read_csv("../output/LFY1_gam_sex_cluster_noSexSamples.csv.gz") %>%
  select(geneID, in_LFY1_cluster)
```

Where is this cluster on the cor.plot?

```{r}
deg.cor %>% left_join(mcconnell_lfy_cluster) %>%
  #arrange(in_LFY1_cluster) %>%
  ggplot(aes(x = LFY1.MR, y = logFC, color = in_LFY1_cluster, alpha = in_LFY1_cluster)) +
  geom_point() +
  scale_alpha_manual(values = c("TRUE" = 0.35, "FALSE" = 0.1)) +
  scale_color_manual(values = c("TRUE" = "red", "FALSE" = "gray")) +
  guides(color = guide_legend(override.aes = list(alpha = 1)))
```


## Compare LFY1 and LFY2 MR in cluster

```{r}
deg.cor %>%
  mutate(upMale = logFC < -1 & FDR < 0.01) %>%
  t.test(LFY1.MR ~ upMale, data = .)

deg.cor %>%
  mutate(upMale = logFC < -1 & FDR < 0.01) %>%
  t.test(LFY2.MR ~ upMale, data = .)
```
```{r}
deg.cor %>%
  t.test(LFY1.MR ~ in_LFY1_cluster, data = .)

deg.cor %>%
  t.test(LFY2.MR ~ in_LFY1_cluster, data = .)
```

```{r}
lfy.network.threshold <- deg.cor %>% filter(in_LFY1_cluster) %>% pull(LFY1.MR) %>% max()

fisher.tmp <- deg.cor %>% mutate(LFY1.network = LFY1.MR <= lfy.network.threshold,
                   LFY2.network = LFY2.MR <= lfy.network.threshold,
                   upMale = logFC < 0 & FDR < 0.01) 

fisher.test(fisher.tmp$upMale, fisher.tmp$LFY1.network, alternative = "greater")
fisher.test(fisher.tmp$upMale, fisher.tmp$LFY2.network, alternative = "greater")

```


## GO enrichment for LFY1 cluster

```{r}
gene2GO.list <- deg.cor %>% 
  rowwise() %>%
  mutate(GOlist = str_split(GO, pattern = " ")) %>%
  ungroup() %>%
  mutate(GOlist = set_names(GOlist, geneID)) %>%
  pull(GOlist)

cluster.genes <- deg.cor %>%
  mutate(cluster.genes = as.numeric(in_LFY1_cluster)) %>% 
  pull(cluster.genes, name = geneID) %>% generics::as.factor()

go.data.cluster <- new("topGOdata", ontology = "BP", description = "In LFY1 Cluster", allGenes = cluster.genes, annot = annFUN.gene2GO, gene2GO = gene2GO.list)

result.cluster <- runTest(go.data.cluster, algorithm = "weight01", statistic = "fisher")

GO.results <- GenTable(go.data.cluster, result.cluster, topNodes = sum(result.cluster@score < 0.05 )) %>%
  dplyr::rename_with( \(x) str_replace(x, "result1|apply.*", "p.value")) %>%
  mutate(p.value = as.numeric(p.value))

GO.results
```

```{r, message=FALSE}
write_csv(GO.results, file = "../output/LFY1_Sex_Cluster_GO_noSexSamples.csv")
```

## Look for overlap with A.t. repro  genes in all male vs hermaphrodite DEG

load gene info

```{r}
Cr_gene_info <- read_csv("../../shared/output/Cr_At_repro_Placket.csv")
```


### Enrichment

For each gene set (up or down), retain genes that are orthologs of A.t. vegetative or one of the reproductive sets and then do Fisher's exact.

```{r}
Cr_gene_info_deg <- deg %>% 
  select(geneID, logFC, FDR) %>%
  inner_join(Cr_gene_info, by=c("geneID"="Crichardii")) %>%
  select(-Athaliana) %>%
  mutate(upHermaphrodite = FDR < 0.01 & logFC > 0,
         upMale = FDR < 0.01 & logFC < 0) %>%
  select(geneID, starts_with("up"), everything(), -FDR, -logFC)

Cr_gene_info_deg


```

Is it ratio of each reproductive category to all others, or to vegetative specific?


```{r}
get.fisher <- function(category, module, t = Cr_gene_info_deg) {
  fisher.test(get(category, t), get(module, t), alternative = "greater")
}

fisher.results <- expand_grid(gene.category = colnames(Cr_gene_info)[-1:-2],
                              module = c("upMale", "upHermaphrodite")) %>%
  mutate(fisher = map2(gene.category, module, get.fisher, Cr_gene_info_deg),
         fisher.tidy = map(fisher, tidy)) %>%
  unnest(fisher.tidy) %>%
  rename(fisher.p.value = p.value) %>%
    mutate(fisher.fdr = p.adjust(fisher.p.value, method = "fdr")) %>%
  arrange(fisher.p.value)

```

```{r}
fisher.results.filter <- fisher.results %>%
  # group_by(module) %>%
  ungroup() %>%
  filter(fisher.fdr < 0.05) %>%
  arrange(module, fisher.fdr) %>%
  select(module, gene.category, estimate, fisher.fdr, everything(), -fisher, -method, -alternative)

#Table 1
write_csv(fisher.results.filter, "../output/male_hermaphrodite_DEG_At_repro_fisher_fdr_0.05.csv")

# Supplemental Table 1
write_csv(fisher.results, "../output/male_hermaphrodite_DEG_At_repro_fisher_all.csv")


fisher.results.filter 
```


## Look for overlap with A.t. repro  genes in male vs hermaphrodite DEG LFY1 cluster

### Enrichment

For each gene set (up or down), retain genes that are orthologs of A.t. vegetative or one of the reproductive sets and then do Fisher's exact.

```{r}
Cr_gene_info_LFY1Cluster <- deg.cor %>% 
  select(geneID, in_LFY1_cluster) %>%
  inner_join(Cr_gene_info, by=c("geneID"="Crichardii")) %>%
  select(-Athaliana)

Cr_gene_info_LFY1Cluster


```

Is it ratio of each reproductive category to all others, or to vegetative specific?


```{r}
get.fisher <- function(category, module, t = Cr_gene_info_LFY1Cluster) {
  fisher.test(get(category, t), get(module, t), alternative = "greater")
}

fisher.results.LFY1.cluster <- expand_grid(gene.category = colnames(Cr_gene_info)[-1:-2],
                              module = "in_LFY1_cluster") %>%
  mutate(fisher = map2(gene.category, module, get.fisher, Cr_gene_info_LFY1Cluster),
         fisher.tidy = map(fisher, tidy)) %>%
  unnest(fisher.tidy) %>%
  rename(fisher.p.value = p.value) %>%
    mutate(fisher.fdr = p.adjust(fisher.p.value, method = "fdr")) %>%
  arrange(fisher.p.value)

write_csv(fisher.results.LFY1.cluster, "../output/male_hermaphrodite_DEG_LFY1_Cluster_At_repro_fisher_all.csv")

```

```{r}
fisher.results.LFY1.cluster.filter <- fisher.results.LFY1.cluster %>%
  # group_by(module) %>%
  ungroup() %>%
  filter(fisher.fdr < 0.05) %>%
  arrange(module, fisher.fdr) %>%
  select(module, gene.category, estimate, fisher.fdr, everything(), -fisher, -method, -alternative)

write_csv(fisher.results.LFY1.cluster.filter, "../output/male_hermaphrodite_DEG_LFY1_Cluster_At_repro_fisher_fdr_0.05.csv")

fisher.results.LFY1.cluster.filter
```