---
title: "12_Sporophyte_A_t_Intersection_Plackett"
author: "Julin Maloof"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(broom)
library(gplots)
library(patchwork)
```

# Using Andy Plackett's ortholog and developmental pathway info.

[Plackett and Catoni](https://www.biorxiv.org/content/10.1101/2025.03.18.643782v1)

## There are at least three things to do with this:

1) Look for enrichment of reproductive genes (sheet 1 for category and sheet 3 for Cr to At info)

2) Update previous GO analysis with Andy's GO set.  Its not hugely different but does have a few more genes annotated and a few more terms

3) Update RNAseq data gene lists with Andy's annotation.

This script will works towards #1

## Details of data From Andy

Breakdown of the sheets:
Arabidopsis gene data: lists of Arabidopsis genes from the various reproductive datasets I culled- each column is a dataset, as labelled. The first column is the vegetative-specific control, and all others were filtered to remove the few genes that overlapped (because no-one quite agrees what 'vegetative specific' is, so I had to take the control on trust)  The source publication is listed for each (and where possible the source file/location within the publication). Where I had to make judgements about what genes from each dataset to include I have listed those too.

Orthofinder orthogroups: created using the Orthofinder program comparing multiple genomes across the land plants (please see the preprint materials and methods for the full list).  Each row represents all related genes from across those genomes, more-or-less at the gene family. You may well be more knowledgeable about this than I, but just in case I would refer to genes from different species within the same orthogroup as 'homologs'

Orthofinder orthologs Cr v At: a more stringent set of relationships output from the Orthofinder analysis, restricted to identifying orthologs between Ceratopteris genes and Arabidopsis genes across the whole Ceratopteris genome. The parent orthogroup is also listed on the far left.  This is what I used to 'translate' Ceratopteris genes into their most direct Arabidopsis relatives (i.e their orthologs, genes descended from the same ancestral gene within the genome). It's not always a one-to-one relationship, and not all Ceratopteris genes have Arabidopsis orthologs (and not all Arabidopsis genes have a Ceratopteris ortholog).  Hopefully you can use this as the basis for assigning relationships between your candidate Ceratopteris genes and Arabidopsis.

Ceratopteris annotation: all the annotation data I was able to glean. Some of this comes from a BLAST search done by the company Novogene when they did our next generation sequencing (columns 'Cr_defline' and 'Cr_genedescription). However, not every Cr gene had annotation based on this, so to try to plug the gaps we looked to the orthologs in the other main land plant models. To do this I appended the annotations from any orthologous genes in Arabidopsis, rice, Selaginella, Physcomitrium, Marchantia, on the same row as the corresponding Cr gene.  This annotation data came from their respective genomes and each is in its own column.  It hasn't filled them all in, but on a surprising number of occasions it has provided an annotation of some sort, even if not Arabidopsis.

Ceratopteris inherited GOterms: to make GOterm enrichment easier, a datet restricted to the GO terms for each Ceratopteris gene extracted from the annotation sheet, including GO terms taken from the orthologous genes from model species.  It covers the whole genome, including Cr genes that had no GO terms inherited.

## Import data
First look at sheet names
```{r}
readxl::excel_sheets("../../shared/input/Ceratopteris_ortholog_data_JM 1.xlsx")
```

```{r}
At_gene_info <- readxl::read_excel("../../shared/input/Ceratopteris_ortholog_data_JM 1.xlsx", 
                                   sheet = "Arabidopsis_gene_data",
                                   skip = 1,
                                   .name_repair = "universal") %>%
  set_names(\(x) str_replace(x, fixed("..."), "."))

At_gene_info_metadata <- At_gene_info[1:2,]

At_gene_info <- At_gene_info[-(1:2), -1]

At_gene_info
```

```{r}
At_gene_info %>%
  summarize(across(everything(), \(x) any(duplicated(na.omit(x)))))
```


```{r}
At_gene_info <- At_gene_info %>%
  pivot_longer(everything(), names_to = "tissue", values_to = "Athaliana") %>%
  na.omit() %>%
  mutate(present = TRUE) %>%
  pivot_wider(names_from = tissue, values_from = present, values_fill = FALSE)

At_gene_info
```

OLD
```{r, eval=FALSE}
Cr_At_orthologs <- readxl::read_excel("../../shared/input/Ceratopteris_ortholog_data_JM 1.xlsx", 
                                      sheet = "Orthofinder_orthologs_Cr_v_At",
                                      .name_repair = "universal") %>%
  set_names(\(x) str_remove(x, "_.*")) 

Cr_At_orthologs
```
NEW
```{r}
Cr_At_orthologs <- read_tsv("../../shared/input/Crichardii_676_v2.1.protein_primaryTranscriptOnly.cleaned__v__Athaliana_167_TAIR10.protein_primaryTranscriptOnly.cleaned.tsv") %>%
  set_names(\(x) str_remove(x, "_.*"))

Cr_At_orthologs
```

### Table of Cr orthlog membership in reproductive groups

```{r}
Cr_At_orthologs <- 
  Cr_At_orthologs %>%
  separate_longer_delim(Crichardii, delim = regex(", ?")) %>%
  separate_longer_delim(Athaliana, delim = regex(", ?")) %>%
  mutate(Athaliana = str_remove(Athaliana, "\\.."),
         Crichardii = str_remove(Crichardii, "\\.p"))

Cr_At_orthologs
```

Update table with Cr genes

```{r}
Cr_gene_info <- inner_join(At_gene_info, Cr_At_orthologs) %>%
  select(Athaliana, Crichardii, everything(), -Orthogroup) 

Cr_gene_info
```



### Load WGCNA network

```{r}
load("../output/WGCNA_gametophyte_all_no4.5DAG.Rdata")

module.assignment <- module.assignment %>%
  mutate(Crichardii = str_remove(geneID, "\\.v2"))

module.assignment

# contains objects: module.assignment, MEs, lcpm.filter, CrLFY1, CrLFY2, MEs.m (A matrix of ME means for plotting), and 
```

```{r}
module.assignment.table <- module.assignment %>%
  select(-geneID) %>%
  mutate(present = TRUE) %>%
  pivot_wider(names_from = module, values_from = present, values_fill = FALSE)
```


## Enrichment

For each module, retain genes that are orthologs of A.t. vegetative or one of the reproductive sets and then do Fisher's exact.

```{r}
Cr_gene_info_module <- module.assignment.table %>% inner_join(Cr_gene_info) %>%
  select(-Athaliana)

Cr_gene_info_module
```

Is it ratio of each reproductive category to all others, or to vegetative specific?


```{r}
get.fisher <- function(category, module, t = Cr_gene_info_module) {
  fisher.test(get(category, t), get(module, t), alternative = "greater")
}

fisher.results <- expand_grid(gene.category = colnames(Cr_gene_info)[-1:-2],
                              module = unique(module.assignment$module)) %>%
  mutate(fisher = map2(gene.category, module, get.fisher, Cr_gene_info_module),
         fisher.tidy = map(fisher, tidy)) %>%
  unnest(fisher.tidy) %>%
  rename(fisher.p.value = p.value) %>%
    mutate(fisher.fdr = p.adjust(fisher.p.value, method = "fdr")) %>%
  arrange(fisher.p.value)

```

```{r}
fisher.results %>%
  # group_by(module) %>%
  ungroup() %>%
  filter(fisher.fdr < 0.05) %>%
  arrange(module, fisher.fdr)
```
## Make a heatmap with module expression levels and enrichment for gene categories...
```{r}
sample.eigen.mean <- sample.eigen.mean %>%
  mutate(label=make.names(label, unique=TRUE)) %>%
  select(label, everything(), - group) %>%
  rename_with(str_remove, pattern = "^ME")
```


```{r}
get_order <- function(x) {x[["labels"]][x[["order"]]]}

sample.order <- sample.eigen.mean %>%
  as.data.frame() %>%
  column_to_rownames("label") %>%
  dist() %>%
  hclust() %>% get_order()

module.order <- sample.eigen.mean %>%
  as.data.frame() %>%
  column_to_rownames("label") %>%
  t() %>%
  dist() %>%
  hclust() %>% get_order()
```


```{r}
ME.pl <- sample.eigen.mean %>%
  pivot_longer(-label, names_to="module", values_to = "expression") %>%
  mutate(module=str_remove(module, "^ME")) %>% 
  ggplot(aes(x=module, y=label, fill=expression)) +
  geom_raster() +
  scale_x_discrete(limits = module.order) +
  scale_y_discrete(limits = sample.order) +
  scale_fill_gradient2(low="blue", high = "red", mid = "white") +
  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=0.5))

ME.pl
```
make a plot for the FDR

```{r}
fisher.results.plot <-
  fisher.results %>%
  filter(fisher.fdr < 0.05) %>%
  select(gene.category, module, estimate)  %>%
  pivot_wider(names_from = module, values_from = estimate)

repro.order <-  fisher.results.plot %>%
  as.data.frame() %>%
  column_to_rownames("gene.category") %>%
  mutate(across(everything(), \(x) replace_na(x, 1))) %>%
  dist() %>%
  hclust() %>% get_order()

fisher.pl <- fisher.results.plot %>%
  pivot_longer(-gene.category, names_to="module", values_to = "enrichment") %>%
  ggplot(aes(x=module, y=gene.category, fill=enrichment)) +
  geom_raster() +
  scale_x_discrete(limits = module.order) +
  scale_y_discrete(limits = repro.order) +
  scale_fill_gradient2(low="blue", high = "red", mid = "white") +
  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=0.5)) +
  ggtitle("CrLFY1: Brown    CrLFY2: Black")

fisher.pl
```

```{r, fig.height = 8, fig.width = 8}
fisher.pl / ME.pl + plot_layout(axes = "collect")
```




