---
title: "14 Attallah EdgeR "
author: "Julin Maloof"
date: "2026-06-10"
output: html_document
---

Goal: Do a differential expression analysis on the Atallah Gametophyte samples

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(edgeR)
library(tidyverse)
library(goseq)
library(GO.db)
library(Biostrings)
library(gplots)
library(ggrepel)
conflicted::conflicts_prefer(dplyr::select, dplyr::filter())
```

```{r}
annotation <- read_delim("../../shared/input/Crichardii_676_v2.1.annotation_info.txt") %>% 
  select(-contains("pacid"), -locusName, -peptideName, geneID = transcriptName)
```


## NCBI counts

### Read the counts and create sample description
```{r}
counts.data.ncbi <- read_delim("../output/ht-seq-counts_ncbi.csv.gz")
head(counts.data.ncbi)
```

```{r}
counts.data.ncbi %>% filter(str_detect(gene_ID, "__"))
```

```{r}
counts.data.ncbi <- counts.data.ncbi %>% filter(str_detect(gene_ID, "__", negate = TRUE))
```


```{r}
sample.description.ncbi <- read_csv("../input/all_srr_info_short.csv")
sample.description.ncbi
```

what is missing?

```{r}
mapped.samples.ncbi <- tibble(Run=colnames(counts.data.ncbi)[-1])
sample.description.ncbi %>% anti_join(mapped.samples.ncbi)
```
callus.  not sure why this got dropped but probably okay.

```{r}
sample.description.ncbi <- sample.description.ncbi %>% semi_join(mapped.samples.ncbi)
```


### Filter for Atallah samples

```{r}
sample.description <- sample.description.ncbi %>% 
  filter(str_detect(description, fixed("4.5 DAG"))) %>%
  mutate(trt = ifelse(str_detect(description, "antheridogen"), "ctrl", "antheridogen")) %>%
  select(sample=Run, trt, description)

counts.data <- counts.data.ncbi[,c("gene_ID", sample.description$sample)]

colnames(counts.data)[-1] == sample.description$sample
```

```{r}
write_csv(counts.data, "../output/counts_unfiltered_Atallah.csv.gz")
```

### more data wrangling

remove low read genes:

```{r}
counts.data <- counts.data[rowSums(counts.data[,-1] > 5) >= 3,]
```

```{r}
counts.matrix <- counts.data %>% select(-gene_ID) %>% as.matrix()
rownames(counts.matrix) <- counts.data$gene_ID

dge.data <- DGEList(counts=counts.matrix, 
                    group=sample.description$trt)
dim(dge.data) 
dge.data <- calcNormFactors(dge.data, method = "TMM")
dge.data$samples # look at the normalization factors
```


## MAKE AN MDS

```{r}
mdsvals <- plotMDS(dge.data, plot = FALSE) # get the MDS values for plotting
```


```{r}
mdsvals2 <- tibble(x=mdsvals$x, y=mdsvals$y, sample=rownames(dge.data$samples)) %>%
  inner_join(sample.description)


mdsvals2 %>% 
  ggplot(aes(x=x, y=y)) +
  theme_bw() +
  geom_point(size=3) +
  geom_text_repel(aes(label=trt), size=2)
```

## Calculate dispersion

```{r}
design <- model.matrix(~trt,data = sample.description) 
colnames(design) <- colnames(design) %>% str_remove("trt")
rownames(design) <- sample.description$sample

design
```

```{r}
#First the overall dispersion
dge.data <- estimateGLMCommonDisp(dge.data,design,verbose = TRUE)

#Then a trended dispersion based on count level
dge.data <- estimateGLMTrendedDisp(dge.data,design)

#And lastly we calculate the gene-wise dispersion, using the prior estimates to "squeeze" the dispersion towards the common dispersion.
dge.data <- estimateGLMTagwiseDisp(dge.data,design)

#We can examine this with a plot
plotBCV(dge.data)
```
## Find DEG

Note: up = plus antheridogen

```{r}
fit <- glmQLFit(dge.data, design)
lrt <- glmQLFTest(fit,coef = "ctrl")
summary(decideTests.DGELRT(lrt,p.value=0.01)) 
```

```{r}
#Extract get DE info for all genes
genes.all <- topTags(lrt, n = Inf, p.value = 1)$table %>%
  rownames_to_column("geneID") %>%
  mutate()
head(genes.all)
```

## Heat maps

Not mean centered
```{r, fig.height = 6}
genes.heatmap <- bind_rows( {
  genes.all %>% 
    filter(logFC > 0) %>% 
    slice_min(PValue, n = 50)
},
{ genes.all %>%
    filter(logFC < 0) %>%
    slice_min(PValue, n = 50)}
  
)

hm <- cpm(dge.data, log = TRUE) %>%
  as_tibble(rownames = "geneID") %>%
    set_names(c("geneID",make.names(as.character(dge.data$samples$group), unique = TRUE))) %>%
  semi_join(genes.heatmap) %>%
  column_to_rownames("geneID") %>%
  as.matrix() %>%
  heatmap.2(trace="none",
            col=colorpanel(15, low="black", mid="#666610", high="#FFFF33"), 
            Colv = FALSE, dendrogram = "row", mar=c(10,8))
  
dev.copy(png, filename = "../output/Atallah_heatmap_tall.png", width=1200, height = 1800)

genes.heatmap.reorder <- genes.heatmap[rev(match(colnames(hm$carpet),genes.heatmap$geneID)),]

write_csv(genes.heatmap.reorder, "../output/Atallah_heatmap_genes_ordered.csv")

dev.off()
```

Mean centered (but not scaled)

```{r}
hm.centered <- cpm(dge.data, log = TRUE) %>%
  as_tibble(rownames = "geneID") %>%
    set_names(c("geneID",make.names(as.character(dge.data$samples$group), unique = TRUE))) %>%
  semi_join(genes.heatmap) %>%
  column_to_rownames("geneID") %>%
  as.matrix() %>%
  t() %>% scale(scale=FALSE) %>% t() %>% # mean-center each gene
  heatmap.2(trace="none",col="bluered", Colv = FALSE, dendrogram = "row", mar=c(10,8))

dev.copy(png, filename = "../output/Atallah_heatmap_tall_centered.png", width=1200, height = 1800)

genes.heatmap.reorder.centered <- genes.heatmap[rev(match(colnames(hm.centered$carpet),genes.heatmap$geneID)),]

write_csv(genes.heatmap.reorder.centered, "../output/Atallah_heatmap_genes_ordered_centered.csv")

dev.off()
```

Add annotation

```{r}
genes.all <- genes.all %>%
  mutate(geneID = str_remove(geneID, "v[1-9]\\.")) %>% left_join(annotation)
```
Add GO term names

```{r, eval=FALSE}
get_GO_definition <- function(go_id) {
  require(GO.db)
  if(is.na(go_id)) return("")
  go_id <- str_split_1(go_id, pattern = " ")
  sapply(go_id, \ (x) {
    term_info <- GOTERM[[x]]
    if (is.null(term_info)) {
      return("")
    }
    Term(term_info)
  }) |>
    str_c(collapse = "; ")
}

genes.all <- genes.all %>% rowwise() %>% mutate(GOterms = get_GO_definition(GO)) %>%
  dplyr::select(geneID:FDR, GO, GOterms, starts_with("Best"), everything())

head(genes.all)

#save to a file
write_csv(genes.all,"../output/DEG_ALL_Atallah.csv.gz")
```

```{r}
genes.all <- read_csv("../output/DEG_ALL_Atallah.csv.gz")
```


## GO term enrichment:

format go list
```{r}
go.list <- genes.all %>% pull(GO) %>% strsplit(" ") %>% set_names(genes.all$geneID)
head(go.list)
```

Get gene lengths
```{r}
gene.lengths <- readRDS("../../shared/input/genelengths.Rds")

gene.lengths <- gene.lengths[genes.all$geneID]
```

vectors of upregulated and downregulated genes

```{r}
genes.up <- genes.all %>% mutate(up = (FDR <= 0.01 & logFC > 0) * 1 ) %>% pull(up, name = geneID)
genes.down <- genes.all %>% mutate(up = (FDR <= 0.01 & logFC < 0) * 1 ) %>% pull(up, name = geneID)
```

checking number of gene with/without GO terms
```{r}
nrow(genes.all) %>% str_c("Total expressed genes: ", .)

genes.all %>% filter(is.na(GO)) %>% nrow() %>% str_c("Genes without GO: ", .)

genes.all %>% filter(!is.na(GO)) %>% nrow() %>% str_c("Genes with GO: ", .)

genes.all %>% filter(!is.na(GO)) %>% nrow() %>% magrittr::divide_by(nrow(genes.all)/100) %>% round(2) %>% str_c("% with GO: ", .)


```


```{r}
#determines if there is bias due to gene length.  The plot shows the relationship.
nullp.result.up <- nullp(DEgenes = genes.up, bias.data = gene.lengths)

#calculate p-values for each GO term
rownames(nullp.result.up) <- names(gene.lengths) #because of a bug in nullp()

GO.out.up <- goseq(pwf = nullp.result.up, gene2cat = go.list, use_genes_without_cat=TRUE) %>%
  filter(ontology == "BP") %>%
  mutate(FDR = p.adjust(over_represented_pvalue, method = "fdr")) %>%
  filter(FDR < 0.05) %>%
  select(category, term, FDR, everything(), -under_represented_pvalue)
 
GO.out.up

write.csv(GO.out.up, "../output/GO_up_antheridogen_Atallah.csv")
```

```{r}
#determines if there is bias due to gene length.  The plot shows the relationship.
nullp.result.down <- nullp(DEgenes = genes.down, bias.data = gene.lengths)

#calculate p-values for each GO term
rownames(nullp.result.down) <- names(gene.lengths) #because of a bug in nullp()

GO.out.down <- goseq(pwf = nullp.result.down, gene2cat = go.list, use_genes_without_cat=TRUE) %>%
  filter(ontology == "BP") %>%
  mutate(FDR = p.adjust(over_represented_pvalue, method = "fdr")) %>%
  filter(FDR < 0.05) %>%
  select(category, term, FDR, everything(), -under_represented_pvalue)
 
GO.out.down

write.csv(GO.out.down, "../output/GO_down_antheridogen_Atallah.csv")

```
