---
title: "17_de_Novo_Motif_Analysis"
output: html_document
---

```{r, message=FALSE}
library(Biostrings)
library(tidyverse)
library(memes)
```

```{r}
setwd(this.path::this.dir()) # for VScode
```

load genome
```{r}
genome <- readDNAStringSet("~/Sequences/ref_genomes/C_richardii_v2.1/assembly/Crichardii_676_v2.0.fa.gz")
genome <- genome[nchar(genome) >= 100000] # only keep reasonably sized scaffolds
```

load gff

```{r}
gff <- rtracklayer::import.gff("~/Sequences/ref_genomes/C_richardii_v2.1/annotation/Crichardii_676_v2.1.gene.gff3.gz") 
gff <- gff[gff$type=="mRNA" & gff$longest==1]
```

```{r}
promoters <- flank(gff, 5000)
# Keep only promoters on scaffolds that are in the genome
promoters <- promoters[seqnames(promoters) %in% names(genome)]
# Set sequence info for promoters to match genome
seqlevels(promoters) <- seqlevels(promoters)[seqlevels(promoters) %in% names(genome)]
suppressWarnings(seqinfo(promoters) <- seqinfo(genome)[seqlevels(promoters)])
# Trim promoters that extend beyond scaffold boundaries
promoters <- suppressWarnings(trim(promoters))
cat("Number of promoters:", length(promoters), "\n")
```

## LFY1 gametophyte sex cluster

```{r}
lfy1.gam.sex.cluster <- read_csv("../output/LFY1_gam_sex_cluster_noSexSamples.csv.gz")
```

### Motif discovery in cluster genes

```{r}
# Filter for genes in the LFY1 cluster
cluster_genes <- lfy1.gam.sex.cluster %>%
  filter(in_LFY1_cluster == TRUE) %>%
  pull(geneID)

cat("Number of genes in LFY1 cluster:", length(cluster_genes), "\n")
```

```{r}
# Match cluster genes to promoters
# First, add gene names to promoters for easier matching
promoters_names_clean <- str_remove(promoters$Name, "\\.v2$")
names(promoters) <- promoters_names_clean

# Remove version suffix if present in geneID
cluster_genes_clean <- str_remove(cluster_genes, "\\.v2$")

# Get promoters for cluster genes by name
cluster_promoters <- promoters[cluster_genes_clean[cluster_genes_clean %in% names(promoters)]]

cat("Number of cluster promoters found:", length(cluster_promoters), "\n")

# Get background promoters (genes NOT in cluster)
background_genes <- lfy1.gam.sex.cluster %>%
  filter(in_LFY1_cluster == FALSE) %>%
  pull(geneID)

background_genes_clean <- str_remove(background_genes, "\\.v2$")
background_promoters <- promoters[background_genes_clean[background_genes_clean %in% names(promoters)]]

cat("Number of background promoters found:", length(background_promoters), "\n")
```

```{r}
# Extract sequences using genome DNAStringSet
cluster_seqs <- genome[cluster_promoters]
names(cluster_seqs) <- names(cluster_promoters)

background_seqs <- genome[background_promoters]
names(background_seqs) <- names(background_promoters)

# Save sequences to fasta for memes
writeXStringSet(cluster_seqs, "../output/LFY1_cluster_promoters.fasta")
writeXStringSet(background_seqs, "../output/LFY1_background_promoters.fasta")
```

```{r}
# Run STREME for discriminative motif discovery
# Using real background promoters instead of shuffled sequences
system.time(
  lfy1_cluster_streme <- runStreme(cluster_seqs, 
                                  control = background_seqs,
                                  nmotifs = 10,
                                  minw = 6,
                                  maxw = 20)
)

lfy1_cluster_streme

write_csv(lfy1_cluster_streme, "../output/lfy1_cluster_denovo_streme.csv")
```

```{r}
# View the motifs
view_tomtom_hits(lfy1_cluster_streme)
```


## WGCNA

### gametophyte samples
```{r}
load("../output/WGCNA_gametophyte_all_no4.5DAG.Rdata")
gmodule.assignment <- module.assignment
```

### LFY1
```{r}
lfy1.gmodule <- gmodule.assignment %>% filter(str_detect(geneID, CrLFY1)) %>% pull(module)

gmodule.assignment %>%
  mutate(geneID = str_remove(geneID, "\\.v2")) %>% 
  right_join(overlaps) %>%
  mutate(module=ifelse(is.na(module), "none", module),
         in.network = module==lfy1.gmodule) %>%
  t.test(count ~ in.network, data = .)
```
```{r}
lfy2.gmodule <- gmodule.assignment %>% filter(str_detect(geneID, CrLFY2)) %>% pull(module)

gmodule.assignment %>%
  mutate(geneID = str_remove(geneID, "\\.v2")) %>% 
  right_join(overlaps) %>%
  mutate(module=ifelse(is.na(module), "none", module),
         in.network = module==lfy2.gmodule) %>%
  t.test(count ~ in.network, data = .)
```

### sporophyte samples
```{r}
load("../output/WGCNA_sporophyte_all.Rdata")
smodule.assignment <- module.assignment
```

### LFY1
```{r}
lfy1.smodule <- smodule.assignment %>% filter(str_detect(geneID, CrLFY1)) %>% pull(module)

smodule.assignment %>%
  mutate(geneID = str_remove(geneID, "\\.v2")) %>% 
  right_join(overlaps) %>%
  mutate(module=ifelse(is.na(module), "none", module),
         in.network = module==lfy1.smodule) %>%
  t.test(count ~ in.network, data = .)
```

```{r}
lfy2.smodule <- smodule.assignment %>% filter(str_detect(geneID, CrLFY2)) %>% pull(module)

smodule.assignment %>%
  mutate(geneID = str_remove(geneID, "\\.v2")) %>% 
  right_join(overlaps) %>%
  mutate(module=ifelse(is.na(module), "none", module),
         in.network = module==lfy2.smodule) %>%
  t.test(count ~ in.network, data = .)
```

## load TGCN networks

## TGCN Gametophyte

### LFY1 Cutoff 6

All genes in submodules
```{r}
read_csv("../output/TGCN_LFY1_gam/results/LFY1_gam_c6_TGCN.csv") %>% 
  rename(geneID = genes) %>% 
  right_join(overlaps) %>%
  mutate(in.network=!is.na(hubGene))  %>%
  t.test(count ~ in.network, data = .)
```
hub genes only
```{r}
read_csv("../output/TGCN_LFY1_gam/results/LFY1_gam_c6_TGCN.csv") %>% 
  select(geneID=hubGene) %>%
  distinct() %>% 
  mutate(in.network=TRUE) %>%
  right_join(overlaps) %>%
  mutate(in.network=!is.na(in.network))  %>%
  t.test(count ~ in.network, data = .)
```

### LFY2 Cutoff 5

Note: this gets more significant as I raise the cutoff (e.g. to 7)

All genes in submodules
```{r}
read_csv("../output/TGCN_LFY2_gam/results/LFY2_gam_c6_TGCN.csv") %>% 
  rename(geneID = genes) %>% 
  right_join(overlaps) %>%
  mutate(in.network=!is.na(hubGene)) %>%
  t.test(count ~ in.network, data = .)

```
hub genes only
```{r}
read_csv("../output/TGCN_LFY2_gam/results/LFY2_gam_c5_TGCN.csv") %>% 
  select(geneID=hubGene) %>%
  distinct() %>% 
  mutate(in.network=TRUE) %>%
  right_join(overlaps) %>%
  mutate(in.network=!is.na(in.network)) %>%
  t.test(count ~ in.network, data = .)
```

## TGCN Sporophyte

### LFY1 Cutoff 5

All genes in submodules
```{r}
read_csv("../output/TGCN_LFY1_sp/results/LFY1_sp_c5_TGCN.csv", show_col_types = FALSE) %>% 
  rename(geneID = genes) %>% 
  right_join(overlaps) %>%
  mutate(in.network=!is.na(hubGene))  %>%
  t.test(count ~ in.network, data = .)
```
hub genes only
```{r}
read_csv("../output/TGCN_LFY1_gam/results/LFY1_gam_c5_TGCN.csv", show_col_types = FALSE) %>% 
  select(geneID=hubGene) %>%
  distinct() %>% 
  mutate(in.network=TRUE) %>%
  right_join(overlaps) %>%
  mutate(in.network=!is.na(in.network))  %>%
  t.test(count ~ in.network, data = .)
```

### LFY2 Cutoff 4

Note: cutoff of 5 or higher is n.s.

All genes in submodules
```{r}
read_csv("../output/TGCN_LFY2_sp/results/LFY2_sp_c4_TGCN.csv", show_col_types = FALSE) %>% 
  rename(geneID = genes) %>% 
  right_join(overlaps) %>%
  mutate(in.network=!is.na(hubGene)) %>%
  t.test(count ~ in.network, data = .)

```
hub genes only
```{r}
read_csv("../output/TGCN_LFY2_sp/results/LFY2_sp_c4_TGCN.csv", show_col_types = FALSE) %>% 
  select(geneID=hubGene) %>%
  distinct() %>% 
  mutate(in.network=TRUE) %>%
  right_join(overlaps) %>%
  mutate(in.network=!is.na(in.network)) %>%
  t.test(count ~ in.network, data = .)
```


Next steps...


Also think about looking at relationship between # of LFY binding sites and correlation


