---
title: "18_Known_Motif_Analysis"
output: html_document
---

```{r, message=FALSE}
library(Biostrings)
library(tidyverse)
library(memes)
library(JASPAR2024)
library(TFBSTools)
library(universalmotif)
```

```{r}
#setwd(this.path::this.dir()) # for VScode
```

load genome
```{r}
genome <- readDNAStringSet("~/Sequences/ref_genomes/C_richardii_v2.1/assembly/Crichardii_676_v2.0.fa.gz")
genome <- genome[nchar(genome) >= 100000] # only keep reasonably sized scaffolds
```

load gff

```{r}
gff <- rtracklayer::import.gff("~/Sequences/ref_genomes/C_richardii_v2.1/annotation/Crichardii_676_v2.1.gene.gff3.gz") 
gff <- gff[gff$type=="mRNA" & gff$longest==1]
```

```{r}
promoters <- flank(gff, 5000)
# Keep only promoters on scaffolds that are in the genome
promoters <- promoters[seqnames(promoters) %in% names(genome)]
# Set sequence info for promoters to match genome
seqlevels(promoters) <- seqlevels(promoters)[seqlevels(promoters) %in% names(genome)]
suppressWarnings(seqinfo(promoters) <- seqinfo(genome)[seqlevels(promoters)])
# Trim promoters that extend beyond scaffold boundaries
promoters <- suppressWarnings(trim(promoters))
cat("Number of promoters:", length(promoters), "\n")
```

## LFY1 gametophyte sex cluster

```{r}
lfy1.gam.sex.cluster <- read_csv("../output/LFY1_gam_sex_cluster_noSexSamples.csv.gz")
```

### Motif discovery in cluster genes

```{r}
# Filter for genes in the LFY1 cluster
cluster_genes <- lfy1.gam.sex.cluster %>%
  filter(in_LFY1_cluster == TRUE) %>%
  pull(geneID)

cat("Number of genes in LFY1 cluster:", length(cluster_genes), "\n")
```

```{r}
# Match cluster genes to promoters
# First, add gene names to promoters for easier matching
promoters_names_clean <- str_remove(promoters$Name, "\\.v2$")
names(promoters) <- promoters_names_clean

# Remove version suffix if present in geneID
cluster_genes_clean <- str_remove(cluster_genes, "\\.v2$")

# Get promoters for cluster genes by name
cluster_promoters <- promoters[cluster_genes_clean[cluster_genes_clean %in% names(promoters)]]

cat("Number of cluster promoters found:", length(cluster_promoters), "\n")

# Get background promoters (genes NOT in cluster)
background_genes <- lfy1.gam.sex.cluster %>%
  filter(in_LFY1_cluster == FALSE) %>%
  pull(geneID)

background_genes_clean <- str_remove(background_genes, "\\.v2$")
background_promoters <- promoters[background_genes_clean[background_genes_clean %in% names(promoters)]]

cat("Number of background promoters found:", length(background_promoters), "\n")
```

```{r}
# Extract sequences using genome DNAStringSet
cluster_seqs <- genome[cluster_promoters]
names(cluster_seqs) <- names(cluster_promoters)

background_seqs <- genome[background_promoters]
names(background_seqs) <- names(background_promoters)
## Get Motifs
```

```{r}
bg_base_freq <- letterFrequency(c(cluster_seqs, background_seqs), letters = c("A", "C", "G", "T"), as.prob = TRUE)
bg_base_freq <- colMeans(bg_base_freq, na.rm = TRUE) / sum(colMeans(bg_base_freq, na.rm = TRUE)) # overcome rounding errors so that it sums to 1
bg_base_freq
```


```{r}
# Get plant transcription factor motifs from JASPAR
# The CORE collection contains high-quality, manually curated motifs
opts <- list(
  collection = "CORE",
  tax_group = "plants",
  all_versions = FALSE  # Get only the latest version of each motif
)

# Access the JASPAR2024 database
JASPAR2024db <- JASPAR2024()@db
pfm_collection <- getMatrixSet(JASPAR2024db, opts)

# Convert PFMatrix to PWMatrix
pwm_collection <- toPWM(pfm_collection, bg=bg_base_freq)

# Convert to universalmotif format for use with memes/FIMO
# This format is required by runFimo()
motif_list <- convert_motifs(pwm_collection, class = "universalmotif-universalmotif")

length(pwm_collection)
```

```{r}
# Determine number of CPUs to use
ncores <- parallel::detectCores() - 1

# Threshold for presence of motif
thresh <- 1e-04

# Split up motifs into groups for parallel processing
motif_groups <- split(
  motif_list,
  cut(seq_along(motif_list), breaks = ncores, labels = FALSE)
)

# Set up parallel processing parameters
bpparam = MulticoreParam(workers = ncores)

# Search all motifs in target promoters using FIMO, use parallel processing
message("Searching target promoters...")
system.time(
  target.matches <- bplapply(
    motif_groups,
    FUN = \(mg) runFimo(sequences = cluster_seqs, 
                        motifs = mg,
                        thresh = thresh,
                        skip_matched_sequence = TRUE),
    BPPARAM = bpparam)
)

# combine the list into a single object
suppressWarnings( target.matches <- do.call(c, unname(target.matches)) )

# Search all motifs at once in non-target promoters  
message("Searching non-target promoters...")
system.time(
  non.target.matches <- bplapply(
    motif_groups,
    FUN = \(mg) runFimo(sequences = background_seqs, 
                        motifs = mg,
                        thresh = thresh,
                        skip_matched_sequence = TRUE),
    BPPARAM = bpparam)
)

# combine the list into a single object
suppressWarnings( non.target.matches <- do.call(c, unname(non.target.matches)) )

# Count unique promoters with matches for each motif
# Note: we are counting presence / absence.  We could also compute average per gene, or sum total
target.counts.df <- target.matches %>%
  as.data.frame() %>%
  group_by(motif_id) %>%
  summarize(target.count = length(unique(seqnames)), .groups = "drop") 

non.target.counts.df <- non.target.matches %>%
  as.data.frame() %>%
  group_by(motif_id) %>%
  summarize(non.target.count = length(unique(seqnames)), .groups = "drop")

# Merge counts (filling in 0s for motifs with no matches)
results.table <- target.counts.df %>%
  full_join(non.target.counts.df, by = "motif_id") %>%
  replace_na(list(target.count = 0, non.target.count = 0))

# Perform Fisher's exact test for each motif
message("Performing statistical tests...")
results.table <- results.table %>%
  rowwise() %>%
  mutate(
    p.value = fisher.test(
      matrix(c(
        target.count,
        length(target.promoters) - target.count,
        non.target.count,
        length(non.target.promoters) - non.target.count
      ), ncol = 2),
      alternative = "greater"
    )$p.value,
    non.target.percent = round(non.target.count / length(non.target.promoters), 3) * 100,
    target.percent = round(target.count / length(target.promoters), 3) * 100
  ) %>%
  ungroup() %>%
  mutate(FDR = p.adjust(p.value, method = "fdr")) %>%
  select(motif = motif_id, non.target.percent, target.percent, FDR, p.value, everything()) %>%
  arrange(p.value)

bpstop(bpparam) # stop the parallel workers

write_csv(results.table, "../output/lfy1_cluster_FIMO_motif_results_trt.csv.gz")
```



## WGCNA

### gametophyte samples
```{r}
load("../output/WGCNA_gametophyte_all_no4.5DAG.Rdata")
gmodule.assignment <- module.assignment
```

### LFY1
```{r}
lfy1.gmodule <- gmodule.assignment %>% filter(str_detect(geneID, CrLFY1)) %>% pull(module)

gmodule.assignment %>%
  mutate(geneID = str_remove(geneID, "\\.v2")) %>% 
  right_join(overlaps) %>%
  mutate(module=ifelse(is.na(module), "none", module),
         in.network = module==lfy1.gmodule) %>%
  t.test(count ~ in.network, data = .)
```
```{r}
lfy2.gmodule <- gmodule.assignment %>% filter(str_detect(geneID, CrLFY2)) %>% pull(module)

gmodule.assignment %>%
  mutate(geneID = str_remove(geneID, "\\.v2")) %>% 
  right_join(overlaps) %>%
  mutate(module=ifelse(is.na(module), "none", module),
         in.network = module==lfy2.gmodule) %>%
  t.test(count ~ in.network, data = .)
```

### sporophyte samples
```{r}
load("../output/WGCNA_sporophyte_all.Rdata")
smodule.assignment <- module.assignment
```

### LFY1
```{r}
lfy1.smodule <- smodule.assignment %>% filter(str_detect(geneID, CrLFY1)) %>% pull(module)

smodule.assignment %>%
  mutate(geneID = str_remove(geneID, "\\.v2")) %>% 
  right_join(overlaps) %>%
  mutate(module=ifelse(is.na(module), "none", module),
         in.network = module==lfy1.smodule) %>%
  t.test(count ~ in.network, data = .)
```

```{r}
lfy2.smodule <- smodule.assignment %>% filter(str_detect(geneID, CrLFY2)) %>% pull(module)

smodule.assignment %>%
  mutate(geneID = str_remove(geneID, "\\.v2")) %>% 
  right_join(overlaps) %>%
  mutate(module=ifelse(is.na(module), "none", module),
         in.network = module==lfy2.smodule) %>%
  t.test(count ~ in.network, data = .)
```

## load TGCN networks

## TGCN Gametophyte

### LFY1 Cutoff 6

All genes in submodules
```{r}
read_csv("../output/TGCN_LFY1_gam/results/LFY1_gam_c6_TGCN.csv") %>% 
  rename(geneID = genes) %>% 
  right_join(overlaps) %>%
  mutate(in.network=!is.na(hubGene))  %>%
  t.test(count ~ in.network, data = .)
```
hub genes only
```{r}
read_csv("../output/TGCN_LFY1_gam/results/LFY1_gam_c6_TGCN.csv") %>% 
  select(geneID=hubGene) %>%
  distinct() %>% 
  mutate(in.network=TRUE) %>%
  right_join(overlaps) %>%
  mutate(in.network=!is.na(in.network))  %>%
  t.test(count ~ in.network, data = .)
```

### LFY2 Cutoff 5

Note: this gets more significant as I raise the cutoff (e.g. to 7)

All genes in submodules
```{r}
read_csv("../output/TGCN_LFY2_gam/results/LFY2_gam_c6_TGCN.csv") %>% 
  rename(geneID = genes) %>% 
  right_join(overlaps) %>%
  mutate(in.network=!is.na(hubGene)) %>%
  t.test(count ~ in.network, data = .)

```
hub genes only
```{r}
read_csv("../output/TGCN_LFY2_gam/results/LFY2_gam_c5_TGCN.csv") %>% 
  select(geneID=hubGene) %>%
  distinct() %>% 
  mutate(in.network=TRUE) %>%
  right_join(overlaps) %>%
  mutate(in.network=!is.na(in.network)) %>%
  t.test(count ~ in.network, data = .)
```

## TGCN Sporophyte

### LFY1 Cutoff 5

All genes in submodules
```{r}
read_csv("../output/TGCN_LFY1_sp/results/LFY1_sp_c5_TGCN.csv", show_col_types = FALSE) %>% 
  rename(geneID = genes) %>% 
  right_join(overlaps) %>%
  mutate(in.network=!is.na(hubGene))  %>%
  t.test(count ~ in.network, data = .)
```
hub genes only
```{r}
read_csv("../output/TGCN_LFY1_gam/results/LFY1_gam_c5_TGCN.csv", show_col_types = FALSE) %>% 
  select(geneID=hubGene) %>%
  distinct() %>% 
  mutate(in.network=TRUE) %>%
  right_join(overlaps) %>%
  mutate(in.network=!is.na(in.network))  %>%
  t.test(count ~ in.network, data = .)
```

### LFY2 Cutoff 4

Note: cutoff of 5 or higher is n.s.

All genes in submodules
```{r}
read_csv("../output/TGCN_LFY2_sp/results/LFY2_sp_c4_TGCN.csv", show_col_types = FALSE) %>% 
  rename(geneID = genes) %>% 
  right_join(overlaps) %>%
  mutate(in.network=!is.na(hubGene)) %>%
  t.test(count ~ in.network, data = .)

```
hub genes only
```{r}
read_csv("../output/TGCN_LFY2_sp/results/LFY2_sp_c4_TGCN.csv", show_col_types = FALSE) %>% 
  select(geneID=hubGene) %>%
  distinct() %>% 
  mutate(in.network=TRUE) %>%
  right_join(overlaps) %>%
  mutate(in.network=!is.na(in.network)) %>%
  t.test(count ~ in.network, data = .)
```


Next steps...


Also think about looking at relationship between # of LFY binding sites and correlation


