---
title: "04_WGCNA"
author: "Julin Maloof"
date: "2025-02-16"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(WGCNA)
library(gplots)
options(stringsAsFactors = FALSE)
```

load sample info
```{r}
sample.description <- read.csv("../input/sample.description.csv")
```


load reads

```{r}
lcpm <- read.csv("../output/log2cpm.csv.gz", row.names = 1, check.names = FALSE)
head(lcpm)
dim(lcpm)
```

Filter for sporophyte samples:

```{r}
sample.description <- sample.description %>% filter(str_detect(tissue, "S5|WYS"))
lcpm <- lcpm[,sample.description$sample]
```


Filter for genes with the highest coefficient of variation

```{r}
CV <- apply(lcpm, 1, \(x) abs(sd(x)/mean(x)))
hist(log10(CV))
```
```{r}
lcpm.filter <- lcpm[CV > quantile(CV, 0.4),]
dim(lcpm.filter)
```

WGCNA wants genes in columns

```{r}
lcpm.filter.t <- t(lcpm.filter)
```


Soft thresholding
```{r}
powers <- c(c(1:10), seq(from = 12, to=20, by=2))
sft <- pickSoftThreshold(lcpm.filter.t, powerVector = powers, verbose = 5,networkType = "signed hybrid", blockSize = 15000)
```

```{r}
sizeGrWindow(9, 5)
par(mfrow = c(1,2))
cex1 <- 0.9
# Scale-free topology fit index as a function of the soft-thresholding power
plot(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     xlab="Soft Threshold (power)",ylab="Scale Free Topology Model Fit,signed R^2",type="n",
     main = paste("Scale independence"))
text(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     labels=powers,cex=cex1,col="red")
# this line corresponds to using an R^2 cut-off of h
abline(h=0.90,col="red")
# Mean connectivity as a function of the soft-thresholding power
plot(sft$fitIndices[,1], sft$fitIndices[,5],
     xlab="Soft Threshold (power)",ylab="Mean Connectivity", type="n",
     main = paste("Mean connectivity"))
text(sft$fitIndices[,1], sft$fitIndices[,5], labels=powers, cex=cex1,col="red")
```
choose 3

```{r}
softPower <- 3
adjacency <- adjacency(lcpm.filter.t, power = softPower, type = "signed hybrid")
# Turn adjacency into topological overlap
TOM <- TOMsimilarity(adjacency, TOMType = "signed");
dissTOM <- 1-TOM
```

```{r}
# Call the hierarchical clustering function
geneTree <- hclust(as.dist(dissTOM), method = "average")
# Plot the resulting clustering tree (dendrogram)
sizeGrWindow(12,9)
plot(geneTree, xlab="", sub="", main = "Gene clustering on TOM-based dissimilarity",
     labels = FALSE, hang = 0.04)
```

define modules

```{r}
# We like large modules, so we set the minimum module size relatively high:
minModuleSize <- 30;
# Module identification using dynamic tree cut:
dynamicMods <- cutreeDynamic(dendro = geneTree, distM = dissTOM,
                             deepSplit <- 2, pamRespectsDendro = FALSE,
                             minClusterSize = minModuleSize);
table(dynamicMods)
```

```{r}
# Convert numeric labels into colors
dynamicColors = labels2colors(dynamicMods)
table(dynamicColors)
# Plot the dendrogram and colors underneath
sizeGrWindow(8,6)
plotDendroAndColors(geneTree, dynamicColors, "Dynamic Tree Cut",
                    dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05,
                    main = "Gene dendrogram and module colors")
```

merge similar modules

```{r}
# Calculate eigengenes
MEList <- moduleEigengenes(lcpm.filter.t, colors = dynamicColors)
MEs <- MEList$eigengenes
# Calculate dissimilarity of module eigengenes
MEDiss <- 1-cor(MEs);
# Cluster module eigengenes
METree <- hclust(as.dist(MEDiss), method = "average");
# Plot the result
sizeGrWindow(7, 6)
plot(METree, main = "Clustering of module eigengenes",
     xlab = "", sub = "")
```

merge with correlation > 0.8
```{r}
MEDissThres = 0.2
# Plot the cut line into the dendrogram
plot(METree, main = "Clustering of module eigengenes",
     xlab = "", sub = "")
abline(h=MEDissThres, col = "red")
# Call an automatic merging function
merge = mergeCloseModules(lcpm.filter.t, dynamicColors, cutHeight = MEDissThres, verbose = 3)
# The merged module colors
mergedColors = merge$colors
# Eigengenes of the new merged modules:
mergedMEs = merge$newMEs
```

compare pre and post merge
```{r}
sizeGrWindow(12, 9)
#pdf(file = "Plots/geneDendro-3.pdf", wi = 9, he = 6)
plotDendroAndColors(geneTree, cbind(dynamicColors, mergedColors),
                    c("Dynamic Tree Cut", "Merged dynamic"),
                    dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05)
#dev.off()
```

```{r}
# Rename to moduleColors
moduleColors = mergedColors
# Construct numerical labels corresponding to the colors
colorOrder = c("grey", standardColors(50));
moduleLabels = match(moduleColors, colorOrder)-1;
MEs = mergedMEs 
table(merge$colors)
length(table(merge$colors))
median(table(merge$colors))

```

## Look at modules

Which module is LFY in?

```{r}
CrLFY1 <- "Ceric.33G031700.2.v2.1" # this is the primary transcript.  There is another but it is expressed at very low levels.

CrLFY2 <- "Ceric.18G076300"

module.assignment <- tibble(geneID=colnames(lcpm.filter.t), module = mergedColors)

module.assignment %>%
  filter(str_detect(geneID, "18G076300|33G031700"))
```
Interesting: they are in different modules(!).  But turquoise is the largest module, so that probably is not very interesting.

```{r}
module.assignment %>% group_by(module) %>% summarize(n_genes = n()) %>% arrange(n_genes)
```

Plot eigengenes

Make sure sample info sheet is in the correct order.
```{r}
rownames(lcpm.filter.t) %>% str_replace_all("\\.", "-") == sample.description$sample
```

```{r}
sample.eigen <- cbind(sample.description, MEs)
sample.eigen
```

```{r, fig.height=6}
sample.eigen.l <- sample.eigen %>%
  mutate(gt_tissue=str_c(base_gt, "-", tissue)) %>%
  pivot_longer(starts_with("ME"), names_to = "ME")

sample.eigen.means <- sample.eigen.l %>%
  group_by(gt_tissue, ME) %>%
  summarise(value = mean(value))

sample.eigen.l %>%
  ggplot(aes(x=gt_tissue, y = value)) +
    geom_point(aes(color = tissue)) +
    geom_line(data=sample.eigen.means, group = 1, lwd=.3) + 
  facet_wrap(~ME, ncol=4) +
  theme(axis.text.x = element_text(angle = 90, hjust=1, vjust=.5)) +
  scale_color_brewer(type="qual", palette = "Set3")
```
A heat map:

```{r, fig.height=7}
MEs.m <- as.matrix(MEs)
heatmap.2(MEs.m, trace="none", cexRow= 0.6, col="bluered")
```



