---
title: "05_GO_WGCNA"
output: html_notebook
---

```{r}
library(tidyverse)
library(topGO)
conflicted::conflicts_prefer(dplyr::select())
#library(clusterProfiler)
```

## for WGCNA with ALL samples
```{r}
load("../output/WGCNA_allSamples.Rdata")
```


```{r}
GOs <- read_delim("../input/Crichardii_676_v2.1.annotation_info.txt")
head(GOs)
tail(GOs)
```

how will do the gene IDs match up?
```{r}
moduleGOs <- module.assignment %>%
  mutate(geneID2 = str_remove(geneID, fixed(".v2"))) %>%
  inner_join(GOs, by = c("geneID2" = "transcriptName")) %>%
  select(geneID, module, GO)

dim(module.assignment)
dim(moduleGOs)
head(moduleGOs)
```
Great.  All genes in module.assignments are retained after the join

Create a gene2GO list needed by topGO:

```{r}
gene2GO.list <- moduleGOs %>% 
  rowwise() %>%
  mutate(GOlist = str_split(GO, pattern = " ")) %>%
  ungroup() %>%
  mutate(GOlist = set_names(GOlist, geneID)) %>%
  pull(GOlist)
```

Create a function to return a go table
```{r}
getGO <- function(moduleID, module.assignment, ontology = "BP")  {
  module.genes <- module.assignment %>%
    mutate(module.genes = as.numeric(module == moduleID)) %>% 
    pull(module.genes, name = geneID) %>% as.factor()
  
  go.data.module <- new("topGOdata", ontology = ontology, description = moduleID, allGenes = module.genes, annot = annFUN.gene2GO, gene2GO = gene2GO.list)
  
  result.module <- runTest(go.data.module, algorithm = "weight01", statistic = "fisher")
  
  GenTable(go.data.module, result.module, topNodes = sum(result.module@score < 0.05 )) %>%
    dplyr::rename_with( \(x) str_replace(x, "result1|apply.*", "p.value"))
}
```


```{r, message=FALSE}
GOresults <- tibble(module=sort(unique(module.assignment$module))) %>%
  mutate(GOtable = map(module, getGO, module.assignment)) %>%
  unnest(cols=GOtable)

write_csv(GOresults, file = "../output/WGCNA_allSamples_GO.csv")

GOresults %>% group_by(module) %>% slice_min(p.value, n=3) %>% write_csv(file = "../output/WGCNA_allSamples_GO_top3.csv")

GOresults
```


## for WGCNA with SPOROPHYTE samples
```{r}
load("../output/WGCNA_sporophyteSamples.Rdata")
```

how will the gene IDs match up?
```{r}
moduleGOs <- module.assignment %>%
  mutate(geneID2 = str_remove(geneID, fixed(".v2"))) %>%
  inner_join(GOs, by = c("geneID2" = "transcriptName")) %>%
  select(geneID, module, GO)

dim(module.assignment)
dim(moduleGOs)
head(moduleGOs)
```
Great.  All genes in module.assignments are retained after the join

Create a gene2GO list needed by topGO:

```{r}
gene2GO.list <- moduleGOs %>% 
  rowwise() %>%
  mutate(GOlist = str_split(GO, pattern = " ")) %>%
  ungroup() %>%
  mutate(GOlist = set_names(GOlist, geneID)) %>%
  pull(GOlist)
```

```{r, message=FALSE}
GOresults <- tibble(module=sort(unique(module.assignment$module))) %>%
  mutate(GOtable = map(module, getGO, module.assignment)) %>%
  unnest(cols=GOtable)

write_csv(GOresults, file = "../output/WGCNA_sporophyteSamples_GO.csv")

GOresults %>% group_by(module) %>% slice_min(p.value, n=3) %>% write_csv(file = "../output/WGCNA_sporophyteSamples_GO_top3.csv")

GOresults
```


## for WGCNA with GAMETOPHYTE samples
```{r}
load("../output/WGCNA_gametophyteSamples.Rdata")
```

how will the gene IDs match up?
```{r}
moduleGOs <- module.assignment %>%
  mutate(geneID2 = str_remove(geneID, fixed(".v2"))) %>%
  inner_join(GOs, by = c("geneID2" = "transcriptName")) %>%
  select(geneID, module, GO)

dim(module.assignment)
dim(moduleGOs)
head(moduleGOs)
```
Great.  All genes in module.assignments are retained after the join

Create a gene2GO list needed by topGO:

```{r}
gene2GO.list <- moduleGOs %>% 
  rowwise() %>%
  mutate(GOlist = str_split(GO, pattern = " ")) %>%
  ungroup() %>%
  mutate(GOlist = set_names(GOlist, geneID)) %>%
  pull(GOlist)
```

```{r, message=FALSE}
GOresults <- tibble(module=sort(unique(module.assignment$module))) %>%
  mutate(GOtable = map(module, getGO, module.assignment)) %>%
  unnest(cols=GOtable)

write_csv(GOresults, file = "../output/WGCNA_gametophyteSamples_GO.csv")

GOresults %>% group_by(module) %>% slice_min(p.value, n=3) %>% write_csv(file = "../output/WGCNA_gametophyteSamples_GO_top3.csv")

GOresults
```
