# Calculate Gene expression

This script is to generate expression data from the .cel files

It will create several different expression datasets:

### From Dataset 319 (Developmental Atlas):
* 319 data sets
  - All Samples
  - Subset 1: 84 samples
  - Subset 2: 72 samples
  - Subset 3: 45 samples chosen to have a diversity of LFY expression levels

I downloaded the Lohman, Schmidt, Weigel AtGenExpress developmenal microarray series described [here](https://www.arabidopsis.org/servlets/TairObject?type=hyb_descr_collection&id=1006710873).  Also see [the paper](http://www.nature.com/ng/journal/v37/n5/full/ng1543.html).  

Samples were downloaded from the [TAIR ftp server](ftp://ftp.arabidopsis.org/home/tair/Microarrays/Datasets/ExpressionSet_ME00319/)

### From Dataset 317 (Floral induction time series)

In this data set, various genotypes are moved from SD to LD and samples are taken at 0, 3, 5, and 7 days after moving.  Essentially a time course of floral induction.

I downloaded the Schmid et al microarray series described [here](https://www.arabidopsis.org/servlets/TairObject?type=expression_set&id=1006710827).  Also see [the paper](http://www.ncbi.nlm.nih.gov/pubmed/14573523?dopt=Abstract).  

Samples were downloaded from the [TAIR ftp server](ftp://ftp.arabidopsis.org/home/tair/Microarrays/Datasets/ExpressionSet_ME00317/)

```{r, echo=TRUE, message=FALSE}
library(here)
library(affy)
#library(gcrma)
library(ath1121501.db) #contains TAIR annotation for the microarray
library(plyr)
library(ggplot2)
library(flashClust)
LFY <- "AT5G61850"
conflicted::conflicts_prefer(here::here)
```

# All samples from ME00319: (will refer to this as set 0)

```{r sample_subset_full}
samples <- read.csv(here("Arabidopsis", "input", "datasets", "ME00319_Schmid", "Sample_Descriptions.csv"), row.names=1)
unique(samples$sample_name)
```

## read in cel files and normalized with rma
```{r readcel_full}
raw0 <- ReadAffy(filenames = paste0(rownames(samples),".gz"),
                 celfile.path = here("Arabidopsis", "input", "datasets",
                                     "ME00319_Schmid"),
                phenoData = samples)
raw0
mas5.0 <- exprs(mas5calls(raw0))
eset0 <- rma(raw0)
eset0.small <- eset0[rowSums(mas5.0 == "P") > 15,] # require "detectable" expression in more than 15 samples
dim(eset0.small)

annotation0 <- select(ath1121501.db,keys = featureNames(eset0.small),columns = c("TAIR","SYMBOL","GENENAME"))

summary(annotation0)
head(annotation0)

#need to deal with duplicates
duplicated <- unique(annotation0$PROBEID[duplicated(annotation0$PROBEID)])

reduced <- ldply(duplicated,function(x) {
  tmp <- annotation0[annotation0$PROBEID==x,]
  tmp[1,"TAIR"] <- paste(unique(na.omit(tmp$TAIR)),sep=".",collapse = ".")
  tmp[1,"SYMBOL"] <- paste(unique(na.omit(tmp$SYMBOL)),sep=".",collapse = ".")
  tmp[1,"GENENAME"] <- paste(unique(na.omit(tmp$GENENAME)),sep=".",collapse = ".")
  tmp[1,]
  })

annotation0 <- annotation0[!annotation0$PROBEID %in% duplicated,]
annotation0 <- rbind(annotation0,reduced)

rownames(annotation0) <- annotation0$PROBEID

annotation.eset0.small <- annotation0[featureNames(eset0.small),] #reoder to match eset order

featureNames(eset0.small) <- make.names(annotation.eset0.small$TAIR,unique=T)
```

I have found that LFY is still expressed at pretty high levels in the _lfy-12_ mutant.  So I will manually adjust the lfy-12 expression level to be the minimum observed.

```{r lfy_Adjust}

gene.expression <- exprs(eset0.small)

qplot(x=colnames(gene.expression),y=gene.expression[LFY,]) +
  ggtitle("LFY expression") +
  theme( axis.text.x  = element_text(angle=90, vjust=0.5, size=10)) +
  xlab("sample") +
  ylab("log2(LFY expression)")

exprs(eset0.small)[LFY,grepl("lfy-12",samples[colnames(gene.expression),"genotype"])] <- min(exprs(eset0.small)[LFY,])

gene.expression <- exprs(eset0.small)

qplot(x=colnames(gene.expression),y=gene.expression[LFY,]) +
  ggtitle("Modified LFY expression") +
  theme( axis.text.x  = element_text(angle=90, vjust=0.5, size=10)) +
  xlab("sample") +
  ylab("log2(LFY expression)")
```

## check cluster sampling.  

We use flashClust to do a quick clustering of samples and learn about the sample relationships.

```{r clustersamples}
exp.norm <- t(exprs(eset0.small))

sampleTree <- flashClust(dist(exp.norm), method = "average");
#pdf(file = "sampleClustering_mean_all.pdf", width = 20, height = 9);
par(cex = 0.6);
par(mar = c(0,4,2,0))
plot(sampleTree, main = "Sample clustering to detect outliers", sub="", xlab="", cex.lab = 1.5,cex.axis = 1.5, cex.main = 2)
plot(sampleTree, main = "Sample clustering to detect outliers", sub="", xlab="", cex.lab = 1.5,cex.axis = 1.5, cex.main = 2,labels=samples$sample_name)
#dev.off()
```

## save the full set

```{r save_full_set}
save(eset0.small, annotation0, annotation.eset0.small, file=here("Arabidopsis", "output", "eset0.RData"))
``` 

# Subset 1: 84 samples


## Choosing samples

I subsetted the original data set to include 84 samples of `r 84/3` sample types.

```{r sample_subset,comment=""}
samples <- read.csv(here("Arabidopsis", "input", "datasets", "ME00319_Schmid", "Sample_Descriptions_Small.csv"), row.names=1)
unique(samples$sample_name)
```

## Read in cel files and normalize with rma
```{r readcel}
raw1 <- ReadAffy(filenames = rownames(samples),
                celfile.path = here("Arabidopsis", "input", "datasets", "ME00319_Schmid"),
                phenoData = samples)
raw1
mas5.1 <- exprs(mas5calls(raw1))
eset1 <- rma(raw1)
eset1.small <- eset1[rowSums(mas5.1 == "P") > 15,] # require "detectable" expression in more than 15 samples
dim(eset1.small)

annotation.eset1.small <- annotation0[featureNames(eset1.small),] #reoder to match eset order

featureNames(eset1.small) <- make.names(annotation.eset1.small$TAIR,unique=T)

save(eset1.small, annotation.eset1.small, file=here("Arabidopsis", "output", "eset1.RData"))
```

# Subset 2: 72 samples

Samples:
```{r sample_subset set2}
samples.2 <- read.csv(here("Arabidopsis", "input", "datasets", "ME00319_Schmid", "Sample_Descriptions_Small2.csv"), row.names=1)
unique(samples.2$sample_name)
```

cel files were read in and normalized with rma
```{r readcel set2}
raw2 <- ReadAffy(filenames = rownames(samples.2),
                  celfile.path = here("Arabidopsis", "input", "datasets", "ME00319_Schmid"),
                  phenoData = samples.2)
raw2
mas5.2 <- exprs(mas5calls(raw2))
eset2 <- rma(raw2)
eset2.small <- eset2[rowSums(mas5.2 == "P") > 15,] # require "detectable" expression in more than 0 samples
dim(eset2.small)

annotation.eset2.small <- annotation0[featureNames(eset2.small),] #reoder to match eset order

featureNames(eset2.small) <- make.names(annotation.eset2.small$TAIR,unique=T)

save(eset2.small, annotation.eset2.small, 
     file = here("Arabidopsis", "output", "eset2.RData"))
```


# Subset 3: LFY expression levels

# Choosing samples

Start with full sample set.  This requires running code in the first part of this script (eset0)

Take a look at _LFY_ transcript levels across these samples.

```{r plot_lfy}
lfy.exp <- exp.norm[sampleTree$order,LFY]
lfy.names <- factor(make.names(samples$sample_name[sampleTree$order],unique=TRUE),
                    levels = make.names(samples$sample_name[sampleTree$order],unique=TRUE))
qplot(lfy.names,y=lfy.exp) + theme(axis.text.x  = element_text(angle=90, vjust=0.5, size=6))
ggsave(here("Arabidopsis", "output", "lfyexp.pdf"), width=14, height=8)

lfy.names <- factor(samples$sample_name[sampleTree$order],
                    levels = unique(samples$sample_name[sampleTree$order]))
qplot(lfy.names,y=lfy.exp) + theme(axis.text.x  = element_text(angle=90, vjust=0.5, size=6))
ggsave(here("Arabidopsis", "output", "lfyexpv2.pdf"), width=10.5, height=8)
```

Based on the plots above, select the following:
```{r choose_samples}
eset3.small <- eset0.small[,pData(eset0.small)$sample %in% c("Col-0_flow12_petal",
                                         "Col-0_shoot_trans",
                                         "Col-0_inflo",
                                         "clv3-7_inflor",
                                         "ufo-1_inflor",
                                         "lfy-12_inflor",
                                         "Col-0_flow15_carp",
                                         "Col-0_flow12_carp",
                                         "clv3-7_flow12",
                                         "Col-0_flow12",
                                         "lfy-12_flow12",
                                         "ufo-1_flow12",
                                         "Col-0_flow9",
                                         "Col-0_flower",
                                         "Col-0_flow10/11")]

annotation.eset3.small <- annotation.eset0.small[featureNames(eset3.small),] #reorder to match eset order

save(eset3.small, annotation.eset3.small, file=here("Arabidopsis", "output", "eset3.RData"))

as.character(unique(pData(eset3.small)$sample))
```

# Dataset 317: Floral induction time series

```{r generate_sample_description, eval=FALSE, echo=FALSE}
#only needs to be done once
sample_description <- data.frame(file_name=dir(here("Arabidopsis", "input", "datasets", "ME00317_Schmid"), pattern="cel"),stringsAsFactors = FALSE)
sample_description <- within(sample_description,{
  sample_name=sub("_[0-9].cel.gz","",file_name)
  genotype=sub("_[0-9]","",sample_name)
  day=sub(".+_","",sample_name)
  })
sample_description
write.csv(sample_description, file=here("Arabidopsis", "input", "datasets", "ME00317_Schmid", "sample_description.csv"), row.names=FALSE)
```


## Choosing samples

I am starting with all samples in the experiment.  These are

```{r sample_subset,comment="",echo=FALSE}
samples <- read.csv(here("Arabidopsis", "input", "datasets", "ME00317_Schmid", "sample_description.csv"), row.names=1)
unique(samples$sample_name)
```

## read in cel files and normalize with rma

```{r readcel}
rawf <- ReadAffy(filenames = rownames(samples),
        celfile.path = here("Arabidopsis", "input", "datasets", "ME00317_Schmid"),
        phenoData = samples,
        compress = TRUE)
rawf
mas5f <- exprs(mas5calls(rawf))
esetf <- rma(rawf)
esetf.small <- esetf[rowSums(mas5f == "P") > 10,] # require "detectable" expression in more than 10 samples
dim(esetf.small)
```



In addition I used the affy mas5 "presence/absence" calls to discard genes that were only present in 10 or fewer samples.  This resulted in keeping about 16,282 genes.

Gene IDs were added from the TAIR database
```{r IDs}
annotationf <- select(ath1121501.db,keys = featureNames(esetf.small),columns = c("TAIR","SYMBOL","GENENAME"))

summary(annotationf)
head(annotationf)

#need to deal with duplicates
duplicated <- unique(annotationf$PROBEID[duplicated(annotationf$PROBEID)])

reduced <- ldply(duplicated,function(x) {
  tmp <- annotationf[annotationf$PROBEID==x,]
  tmp[1,"TAIR"] <- paste(unique(na.omit(tmp$TAIR)),sep=".",collapse = ".")
  tmp[1,"SYMBOL"] <- paste(unique(na.omit(tmp$SYMBOL)),sep=".",collapse = ".")
  tmp[1,"GENENAME"] <- paste(unique(na.omit(tmp$GENENAME)),sep=".",collapse = ".")
  tmp[1,]
  })

annotationf <- annotationf[!annotationf$PROBEID %in% duplicated,]
annotationf <- rbind(annotationf,reduced)

rownames(annotationf) <- annotationf$PROBEID

annotation.esetf.small <- annotationf[featureNames(esetf.small),] #reoder to match eset order

featureNames(esetf.small) <- make.names(annotation.esetf.small$TAIR,unique=T)
```


## Adjust lfy-1

I realized that _lfy-12_ does not cause a reduction in _LFY_ transcript level and this could mess up the correlations/networks, so I will reassign the _LFY_ transcript level in _lfy-12_ mutants to the lowest observed _LFY_ level in any sample

```{r print_expression, echo=FALSE}

gene.expression <- exprs(esetf.small)

qplot(x=colnames(gene.expression),y=gene.expression[LFY,]) +
  ggtitle("LFY expression") +
  theme( axis.text.x  = element_text(angle=90, vjust=0.5, size=10)) +
  xlab("sample") +
  ylab("log2(LFY expression)")

exprs(esetf.small)[LFY,grepl("LFY12",samples[colnames(gene.expression),"genotype"])] <- min(exprs(esetf.small)[LFY,])

gene.expression <- exprs(esetf.small)

qplot(x=colnames(gene.expression),y=gene.expression[LFY,]) +
  ggtitle("Modified LFY expression") +
  theme( axis.text.x  = element_text(angle=90, vjust=0.5, size=10)) +
  xlab("sample") +
  ylab("log2(LFY expression)")

```

## save the 317 set

```{r save_317_set}
save(esetf.small, annotationf, annotation.esetf.small, file=here("Arabidopsis", "output", "esetf.RData"))
```
