---
title: "02 - Mutual Rank Analysis"
author: "Julin Maloof"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

Want to generate mutual rank (MR) for the ME00319 and ME00317 datasets.

## Required Libraries

```{r libraries}
library(tidyverse)
library(igraph)
library(snowfall)
library(knitr)
library(GOstats)
library(org.At.tair.db)
library(pheatmap)
library(here)
```

## Set Constants

```{r constants}
LFY <- "AT5G61850"  # LEAFY gene ID
```

## Load Known LFY Targets

These are targets from Winters et al.
```{r load_targets}
load(here("Arabidopsis", "input", "LFY_targets.RData"))
```

Load _a priori_ targets from Hannah

```{r load_hannah_targets}
apriori.targets <- read_csv(here("Arabidopsis", "input", "LFY GRN candidates.csv"), name_repair = "universal") %>% 
mutate(Gene.name = ifelse(Gene.name == "*", AGI, Gene.name),
        AGI = toupper(AGI),
        Cr.match = Cr.ID,
        Cr.ID = str_extract(Cr.match, "Ceric\\.[0-9]{2}G[0-9]{6}\\.[0-9]")) %>% 
        dplyr::select(everything(), Cr.match)

apriori.targets %>% print(n=100)
```

# Mutual Rank Analysis

## Helper Functions for MR Analysis

```{r mr_functions}
# Function to calculate mutual ranks
calc_mr <- function(eset, outfile, LFYonly = TRUE) {
  # There isn't enough memory on my laptop to return the full resutls on multiple sets,
  # so we save the results to a file if outfile is provided and give an option to return
  # only LFY-related results.

  eset <- exprs(eset)  # Extract expression data

 # eset <- eset[c(match(c(LFY, apriori.targets$AGI), rownames(eset)),  1:100),] # For testing, use only LFY, a priori candidates, and first 100 genes

  # Calculate pairwise correlations
  gene.cor <- cor(t(eset))

  diag(gene.cor) <- 0
  
  # Rank correlations for each gene
  gene.rank <- apply(gene.cor, 2, function(x) rank(-abs(x)))
  
  # Calculate mutual rank as geometric mean of ranks
  gene.mr <- sqrt(gene.rank * t(gene.rank))

  if(!is.null(outfile)) {
    saveRDS(list(cor = gene.cor, mr = gene.mr, rank = gene.rank), file = outfile)
  }

  if(LFYonly) {
    # extract LFY columns
    lfy.cor <- gene.cor[, LFY]
    lfy.rank <- gene.rank[, LFY]
    lfy.mr <- gene.mr[, LFY]
    return(tibble(AGI=names(lfy.cor), lfy.cor, lfy.rank, lfy.mr))
  } else {
    return(list(cor = gene.cor, rank = gene.rank, mr = gene.mr))
  }
}
```



## Load Data Subsets

```{r load_subsets}
load(here("Arabidopsis", "output", "eset0.Rdata"))

load(here("Arabidopsis", "output", "eset1.RData"))

load(here("Arabidopsis", "output", "eset2.RData"))

load(here("Arabidopsis", "output", "eset3.RData"))

load(here("Arabidopsis", "output", "esetf.RData"))
```


## Get MR for each subset

```{r analyze_subsets}

Vsize <- mem.maxVSize()

mem.maxVSize(max(Vsize, 32000)) 

results <- tibble(name=ls(".GlobalEnv", pattern = "^eset.\\.")) %>%
  mutate(MR = map(name, ~ {
    calc_mr(get(.x),
    outfile = here("Arabidopsis", "output", paste0(tools::file_path_sans_ext(.x), "_MR.RData")))
  }, 
  .progress = TRUE)) 

saveRDS(results, file = here("Arabidopsis", "output", "MR_LFY_results.Rds"))

mem.maxVSize(Vsize) # return to original memory size
```

```{r calculate_distance}
results <- results %>% 
  mutate(target.distance = map(MR, inner_join, apriori.targets), 
         set.size = nrow(MR))

saveRDS(results, file = here("Arabidopsis", "output", "MR_LFY_results.Rds"))

results %>% unnest(target.distance)
```


```{r summarize_distance}
results %>% 
  unnest(target.distance) %>% 
  group_by(name) %>%
  summarize(across(starts_with("lfy"),
    .fns = list(
      mean = \(x) mean(abs(x), na.rm=TRUE),
      median = \(x) median(abs(x), na.na.rm = TRUE)
      )))

```

```{r plot_distance}
results %>% unnest(target.distance) %>% 
  pivot_longer(starts_with("lfy"), values_to = "score", names_to = "metric") %>% 
  ggplot(aes(x=Gene.name, y = score, fill = name)) +
  geom_col(position = "dodge") +
  facet_wrap(~metric, ncol=1, scales = "free_y") +
  scale_fill_brewer(type="qual", palette = 3)
```


focus on the 3 best ones
```{r plot_distance}
results %>% 
  filter(str_detect(name, "2|3|f")) %>%
  unnest(target.distance) %>% 
  pivot_longer(starts_with("lfy"), values_to = "score", names_to = "metric") %>% 
  ggplot(aes(x=Gene.name, y = score, fill = name)) +
  geom_col(position = "dodge") +
  facet_wrap(~metric, ncol=1, scales = "free_y") +
  scale_fill_brewer(type="qual", palette = 3)
```


NOTE: MAYBE I SHOULD REDO THIS WITHOUT THE LFY-12 MUTANT?

I AM HERE


# Unused Functions

```{r}
# Function to extract subgraph at a specified MR cutoff
get_mr_subgraph <- function(mr.cutoff, mr.matrix, annotation, neighborhood = LFY, order = 1) {
  # Function to extract LFY graph at a specificed MR cutoff
  gene.mr.tmp <- mr.matrix
  gene.mr.tmp[abs(gene.mr.tmp) > mr.cutoff] <- 0
  gene.mr.tmp[is.na(gene.mr.tmp)] <- 0 # important! otherwise vertices with NA edges are connected
  
  gene.graph <- graph.adjacency(adjmatrix = gene.mr.tmp,
                                mode = "undirected",
                                weighted = "mr",
                                diag = FALSE)
  
  sub.graph <- graph.neighborhood(gene.graph, order = order, nodes = neighborhood)[[1]]
  
  V(sub.graph)$color <- "lightblue"
  V(sub.graph)[neighborhood]$color <- "red"
  V(sub.graph)$gene <- annotation$SYMBOL[match(V(sub.graph)$name, annotation$TAIR)]
  
  list(cutoff = mr.cutoff, graph = sub.graph)
}

# Function to evaluate overlap with known targets
target_overlap <- function(lfy_graph, targets, inf_targets, seedling_targets) {
  vnames <- V(lfy_graph$graph)$name
  
  # Some IDs may contain dots, so we split them
  vnames <- unlist(strsplit(vnames, split = ".", fixed = TRUE))
  
  c(LFY_nodes = length(vnames) - 1,
    Known_Inf_targets = length(inf_targets) - 1,
    Known_Seedling_targets = length(seedling_targets),
    Known_targets = length(targets) - 1,
    Intersection = length(intersect(vnames, targets)) - 1,
    PercentNetworkIsKnownInfTarget = 100 * ((length(intersect(vnames, inf_targets)) - 1) / (length(vnames) - 1)),
    PercentNetworkIsKnownSeedlingTarget = 100 * ((length(intersect(vnames, seedling_targets)) - 1) / (length(vnames) - 1)),
    PercentNetworkIsKnownAnyTarget = 100 * ((length(intersect(vnames, targets)) - 1) / (length(vnames) - 1)),
    PercentKnownInfTargetsInNetwork = 100 * ((length(intersect(vnames, inf_targets)) - 1) / (length(inf_targets) - 1)),
    PercentKnownSeedlingTargetsInNetwork = 100 * ((length(intersect(vnames, seedling_targets)) - 1) / (length(seedling_targets))))
}

# Function to perform GO enrichment analysis
GO_enrichment <- function(lfy_graph, universe) {
  # Extract gene names from graph
  vnames <- V(lfy_graph$graph)$name
  vnames <- unlist(strsplit(vnames, split = ".", fixed = TRUE))
  
  # Remove LFY itself
  vnames <- setdiff(vnames, LFY)
  
  # Create GOHyperGParams object
  params <- new("GOHyperGParams",
                geneIds = vnames,
                universeGeneIds = universe,
                annotation = "org.At.tair.db",
                ontology = "BP",
                pvalueCutoff = 0.01,
                conditional = FALSE,
                testDirection = "over")
  
  # Perform hypergeometric test
  hyper_test <- tryCatch({
    hyperGTest(params)
  }, error = function(e) {
    message("GO enrichment failed: ", e$message)
    return(NULL)
  })
  
  # If successful, format the results
  if (!is.null(hyper_test)) {
    go_table <- summary(hyper_test)
    go_table <- go_table[order(go_table$Pvalue), ]
    return(go_table[1:min(10, nrow(go_table)), ])
  } else {
    return(NULL)
  }
}
```

# Analyze Data Subsets

DON'T USE THIS
```{r analyze_function, eval=FALSE}
# Create a consolidated function for analyzing any subset
analyze_subset <- function(eset_data, subset_name) {
  # retreive expression data
  exp.norm <- exprs(eset_data)
  
  # Calculate mutual ranks
  mr.results <- calc_mr(exp.norm)
  gene.mr <- mr.results$mr
  
  # Get LFY networks at different MR thresholds
  cutoffs <- c(10, 20, 30, 50, 80, 100, 130, 210)
  lfy.mr.graphs <- lapply(cutoffs, get_mr_subgraph, mr.matrix = gene.mr, 
                          annotation = annotation, neighborhood = LFY)
  names(lfy.mr.graphs) <- sapply(lfy.mr.graphs, function(x) paste("lfy.mr.graph", x$cutoff, sep = "."))
  
  # Compare to known targets
  overlap.summary <- lapply(lfy.mr.graphs, target_overlap, 
                           targets = targets, inf.targets = inf_targets, 
                           seedling.targets = seedling_targets)
  overlap.summary <- do.call(cbind, overlap.summary)
  colnames(overlap.summary) <- names(lfy.mr.graphs)
  
  # GO enrichment for each network
  go.results <- list()
  for (i in seq_along(lfy.mr.graphs)) {
    # Create universe of gene IDs
    universe <- rownames(exp.norm)
    universe <- unique(unlist(strsplit(universe, ".", fixed = TRUE)))
    
    go.results[[names(lfy.mr.graphs)[i]]] <- GO_enrichment(lfy.mr.graphs[[i]], universe)
  }
  
  # Display results
  cat(paste0("\n## Results for ", subset_name, "\n"))
  
  # Display overlap summary
  print(kable(t(overlap.summary), 
        caption = paste0("Overlap of MR networks with known LFY targets (", subset_name, ")")))
  
  # Display GO enrichment for MR = 80
  cat(paste0("\nGO Enrichment for MR = 80 (", subset_name, "):\n"))
  
  if (!is.null(go.results$lfy.mr.graph.80) && nrow(go.results$lfy.mr.graph.80) > 0) {
    print(kable(go.results$lfy.mr.graph.80, 
          caption = paste0("GO Enrichment for MR = 80 network (", subset_name, ")")))
  } else {
    cat("No significant GO enrichment found.\n")
  }
  
  # Return results
  return(list(graphs = lfy.mr.graphs, 
              overlap = overlap.summary, 
              go.enrichment = go.results,
              cutoffs = cutoffs))
}
```

# Visualize MR Networks

We'll visualize the network with the best enrichment for each subset.

```{r visualize_networks, fig.width=10, fig.height=8, eval = FALSE}
# Function to visualize network
visualize_mr_network <- function(lfy_graph, title) {
  plot(lfy_graph$graph, 
       layout = layout_with_fr(lfy_graph$graph),
       vertex.size = 8,
       vertex.label = V(lfy_graph$graph)$gene,
       vertex.label.cex = 0.7,
       edge.width = 1/E(lfy_graph$graph)$mr * 5,
       main = paste0(title, ", MR cutoff = ", lfy_graph$cutoff))
}

# Determine how many subsets we have
subset.names <- names(subset.results)
num.subsets <- length(subset.names)

if (num.subsets > 0) {
  # Arrange the plot layout based on number of subsets
  par(mfrow = c(1, min(3, num.subsets)))
  
  # Visualize each available network
  for (i in seq_along(subset.names)) {
    subset.name <- subset.names[i]
    display.name <- gsub("subset", "Subset ", subset.name)
    
    # Check if this subset has results and the MR=80 network
    if (length(subset.results[[subset.name]]) > 0 && 
        "lfy.mr.graph.80" %in% names(subset.results[[subset.name]]$graphs)) {
      visualize_mr_network(subset.results[[subset.name]]$graphs$lfy.mr.graph.80, display.name)
    }
  }
}
```

# Compare Results Across Subsets

```{r compare_results, eval = FALSE}
# Function to extract comparison metrics
extract_comparison_metrics <- function(results_list, cutoff = 80) {
  if (length(results_list) == 0) {
    return(NULL)
  }
  
  graph.name <- paste0("lfy.mr.graph.", cutoff)
  overlap <- results_list$overlap[, graph.name]
  
  go.enrichment <- results_list$go.enrichment[[graph.name]]
  if (!is.null(go.enrichment) && nrow(go.enrichment) > 0) {
    top.go <- go.enrichment$Term[1]
    top.go.pvalue <- go.enrichment$Pvalue[1]
  } else {
    top.go <- "None"
    top.go.pvalue <- NA
  }
  
  c(LFY_Nodes = overlap["LFY_nodes"],
    Percent_Known_Targets = overlap["PercentNetworkIsKnownAnyTarget"],
    Top_GO_Term = top.go,
    Top_GO_Pvalue = top.go.pvalue)
}

# Get list of subset names and create comparison table
subset.names <- names(subset.results)
display.names <- gsub("subset", "Subset ", subset.names)

# Initialize comparison table with the right dimensions
comparison.table <- data.frame(
  Subset = display.names,
  LFY_Nodes = NA,
  Percent_Known_Targets = NA,
  Top_GO_Term = NA,
  Top_GO_Pvalue = NA,
  stringsAsFactors = FALSE
)

# Fill in the comparison table
for (i in seq_along(subset.names)) {
  subset.name <- subset.names[i]
  
  # Extract metrics if results are available
  if (length(subset.results[[subset.name]]) > 0) {
    metrics <- extract_comparison_metrics(subset.results[[subset.name]])
    if (!is.null(metrics)) {
      comparison.table$LFY_Nodes[i] <- metrics["LFY_Nodes"]
      comparison.table$Percent_Known_Targets[i] <- metrics["Percent_Known_Targets"]
      comparison.table$Top_GO_Term[i] <- metrics["Top_GO_Term"]
      comparison.table$Top_GO_Pvalue[i] <- metrics["Top_GO_Pvalue"]
    }
  }
}

# Display comparison table
kable(comparison.table, 
      caption = "Comparison of MR = 80 Networks Across Subsets",
      row.names = FALSE)
```

# Save Results

```{r save_results, eval = FALSE}
# Save all results
all.results <- list(
  subset.results = subset.results,
  comparison = comparison.table
)

save(all.results, file = here("Arabidopsis", "output", "MR_ME00319_results.RData"))

# Save individual subset results for easier access
for (subset.name in names(subset.results)) {
  if (length(subset.results[[subset.name]]) > 0) {
    subset.data <- subset.results[[subset.name]]
    save.file <- here("Arabidopsis", "output", paste0("MR_ME00319_", subset.name, ".RData"))
    save(subset.data, file = save.file)
  }
}
```
