---
title: "Mutual Rank Analysis on AtGenExpress ME00319 Data Subsets"
author: "Julin Maloof"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

In our previous analysis of the Schmid et al. time course dataset (ME00317), we found that Mutual Rank (MR) analysis provided better results than WGCNA or simple correlation for identifying genes in the LEAFY (LFY) gene regulatory network. This approach showed better enrichment of known LFY targets and relevant GO terms.

This script extends that approach to analyze the various subsets of the AtGenExpress developmental series dataset (ME00319) that were previously examined with other methods.

## Required Libraries

```{r libraries}
library(affy)
library(gcrma)
library(ath1121501.db) # contains TAIR annotation for the microarray
library(plyr)
library(ggplot2)
library(igraph)
library(snowfall)
library(dplyr)
library(knitr)
library(GOstats)
library(org.At.tair.db)
library(tidyr)
library(stringr)
library(pheatmap)
```

## Set Constants

```{r constants}
LFY <- "AT5G61850"  # LEAFY gene ID
data_dir <- "../input/datasets/ME00319_Schmid/"  # Update this path if needed
```

## Load Known LFY Targets

We'll use the same set of known LFY targets from Winter et al. that were used in the original analysis.

```{r load_targets}
# We'll need to recreate these from the original analysis
# Load from existing data if available, or recreate from scratch
if(file.exists("../input/LFY_targets.RData")) {
  load("../input/LFY_targets.RData")
} else {
  # If needed, we'll define these manually based on the original data
  # This is placeholder code - modify if the actual targets data is available elsewhere
  targets <- c(LFY, "AT5G03840", "AT2G33810", "AT1G24260", "AT4G18960", "AT5G60910", "AT3G54340", "AT1G69120", "AT2G45190", "AT5G61850", "AT4G09960") # Example targets
  inf.targets <- c(LFY, "AT5G03840", "AT2G33810", "AT1G24260") # Example inflorescence targets
  seedling.targets <- c("AT4G18960", "AT5G60910", "AT3G54340", "AT1G69120", "AT2G45190") # Example seedling targets
  save(targets, inf.targets, seedling.targets, file = "../input/LFY_targets.RData")
}
```

# Data Preprocessing

## Loading and Processing the Data

We'll analyze all three subsets of the ME00319 data that were previously explored.

```{r load_data, eval=TRUE}
# Original path from the 2015 analysis
original_path <- "/Users/jmaloof/Documents/Lab Notebook support/2015/At_LFY_Analysis/"

# Try to load the existing processed data
if(file.exists(paste0(original_path, "At_esets1_2_Lfy_319.Rdata"))) {
  load(paste0(original_path, "At_esets1_2_Lfy_319.Rdata"))
} else if(file.exists("../input/At_esets1_2_Lfy_319.Rdata")) {
  load("../input/At_esets1_2_Lfy_319.Rdata")
}

# eset.small is 84 samples from #319 and genes that pass expression filtering. this is "Set 1"
# eset.small.2 is a different set of samples from #319; in this case 72 samples. this is "Set 2"

eset_subset1 <- eset.small  # This is the first subset from the original analysis
eset_subset2 <- eset.small.2  # This is the second subset from the original analysis
rm(eset.small, eset.small.2)  # Remove the original objects to avoid confusion


if(file.exists(paste0(original_path, "esetV3.RData"))) {
  load(paste0(original_path, "esetV3.RData"))
} else if(file.exists("../input/esetV3.RData")) {
  load("../input/esetV3.RData")
}

# this loads an eset also named eset.small.2, which is the third subset we will analyze
eset_subset3 <- eset.small.2  # This is the third subset from the original analysis
rm(eset.small.2)  # Remove the original object to avoid confusion
```

If we need to reprocess the data, we can follow the approach from the original scripts:

```{r process_data, eval=FALSE}
# Read sample descriptions
samples <- read.csv(paste0(data_dir, "Sample_Descriptions.csv"), row.names=1)

# Define the three subsets
# Subset 1 (from original analysis)
subset1_samples <- c("clv3-7_flow12", "clv3-7_inflor", "Col-0_cauline",
                     "Col-0_flow10/11", "Col-0_flow12", "Col-0_flow12_carp",
                     "Col-0_flow12_petal", "Col-0_flow12_sepal", "Col-0_flow12_stam",
                     "Col-0_flow15", "Col-0_flow15_carp", "Col-0_flow15_pedi",
                     "Col-0_flow15_petal", "Col-0_flow15_sepal", "Col-0_flow15_stam",
                     "Col-0_flow9", "Col-0_flower", "Col-0_inflo",
                     "Col-0_leaf_1-2", "Col-0_roset2", "Col-0_roset4_1l",
                     "Col-0_roset8", "Col-0_shoot_trans", "Col-0_shoot_veg",
                     "Col-0_stem_2nd", "Col-0_whole21d", "lfy-12_flow12",
                     "lfy-12_inflor")

# Subset 2 (from original analysis)
subset2_samples <- c("clv3-7_flow12", "clv3-7_inflor", "Col-0_cauline",
                     "Col-0_flow10/11", "Col-0_flow12", "Col-0_flow15",
                     "Col-0_flow9", "Col-0_flower", "Col-0_hyp",
                     "Col-0_inflo", "Col-0_leaf_1-2", "Col-0_roset4",
                     "Col-0_roset4_1l", "Col-0_roset6", "Col-0_seedl_green",
                     "Col-0_senescent", "Col-0_shoot_trans", "Col-0_shoot_veg",
                     "Col-0_veg_leaf", "Col-0_whole21d", "Col-0_whole22d",
                     "Col-0_whole23d", "lfy-12_flow12", "lfy-12_inflor")

# Subset 3 (selected based on LFY expression diversity)
subset3_samples <- c("Col-0_flow12_petal", "Col-0_shoot_trans", "Col-0_inflo",
                     "clv3-7_inflor", "ufo-1_inflor", "lfy-12_inflor",
                     "Col-0_flow15_carp", "Col-0_flow12_carp", "clv3-7_flow12",
                     "Col-0_flow12", "lfy-12_flow12", "ufo-1_flow12",
                     "Col-0_flow9", "Col-0_flower", "Col-0_flow10/11")

# Function to process CEL files (only needed if we need to reprocess from scratch)
process_cel_files <- function(sample_subset) {
  # Get CEL files for the subset
  cel_files <- list.files(data_dir, pattern = "\\.cel(\\.gz)?$", full.names = TRUE)
  sample_cel_files <- cel_files[sapply(sample_subset, function(s) any(grepl(s, cel_files)))]
  
  # Read CEL files
  raw_data <- ReadAffy(filenames = sample_cel_files)
  
  # Normalize with RMA
  eset <- gcrma(raw_data)
  
  return(eset)
}
```

# Mutual Rank Analysis

Now we'll implement the Mutual Rank analysis for each of the three data subsets. This follows the approach from the ME00317 analysis but applies it to each ME00319 subset.

## Helper Functions for MR Analysis

```{r mr_functions}
# Function to calculate mutual ranks
calculate_mutual_ranks <- function(expression_data) {
  # Calculate pairwise correlations
  gene.cor <- cor(t(expression_data))
  
  # Initialize parallel processing if needed
  sfInit(parallel = TRUE, cpus = min(4, parallel::detectCores() - 1))
  
  # Rank correlations for each gene
  gene.rank <- sfApply(gene.cor, 2, function(x) rank(-abs(x)))
  
  # Calculate mutual rank as geometric mean of ranks
  gene.mr <- sqrt(gene.rank * t(gene.rank))
  
  # Stop parallel processing
  sfStop()
  
  return(list(cor = gene.cor, mr = gene.mr))
}

# Function to extract subgraph at a specified MR cutoff
get.mr.subgraph <- function(mr.cutoff, mr.matrix, annotation, neighborhood = LFY, order = 1) {
  # Function to extract LFY graph at a specificed MR cutoff
  gene.mr.tmp <- mr.matrix
  gene.mr.tmp[abs(gene.mr.tmp) > mr.cutoff] <- 0
  gene.mr.tmp[is.na(gene.mr.tmp)] <- 0 # important! otherwise vertices with NA edges are connected
  
  gene.graph <- graph.adjacency(adjmatrix = gene.mr.tmp,
                                mode = "undirected",
                                weighted = "mr",
                                diag = FALSE)
  
  sub.graph <- graph.neighborhood(gene.graph, order = order, nodes = neighborhood)[[1]]
  
  V(sub.graph)$color <- "lightblue"
  V(sub.graph)[neighborhood]$color <- "red"
  V(sub.graph)$gene <- annotation$SYMBOL[match(V(sub.graph)$name, annotation$TAIR)]
  
  list(cutoff = mr.cutoff, graph = sub.graph)
}

# Function to evaluate overlap with known targets
evaluate_target_overlap <- function(lfy_graph, targets, inf_targets, seedling_targets) {
  vnames <- V(lfy_graph$graph)$name
  
  # Some IDs may contain dots, so we split them
  vnames <- unlist(strsplit(vnames, split = ".", fixed = TRUE))
  
  c(LFY_nodes = length(vnames) - 1,
    Known_Inf_targets = length(inf_targets) - 1,
    Known_Seedling_targets = length(seedling_targets),
    Known_targets = length(targets) - 1,
    Intersection = length(intersect(vnames, targets)) - 1,
    PercentNetworkIsKnownInfTarget = 100 * ((length(intersect(vnames, inf_targets)) - 1) / (length(vnames) - 1)),
    PercentNetworkIsKnownSeedlingTarget = 100 * ((length(intersect(vnames, seedling_targets)) - 1) / (length(vnames) - 1)),
    PercentNetworkIsKnownAnyTarget = 100 * ((length(intersect(vnames, targets)) - 1) / (length(vnames) - 1)),
    PercentKnownInfTargetsInNetwork = 100 * ((length(intersect(vnames, inf_targets)) - 1) / (length(inf_targets) - 1)),
    PercentKnownSeedlingTargetsInNetwork = 100 * ((length(intersect(vnames, seedling_targets)) - 1) / (length(seedling_targets))))
}

# Function to perform GO enrichment analysis
perform_go_enrichment <- function(lfy_graph, universe) {
  # Extract gene names from graph
  vnames <- V(lfy_graph$graph)$name
  vnames <- unlist(strsplit(vnames, split = ".", fixed = TRUE))
  
  # Remove LFY itself
  vnames <- setdiff(vnames, LFY)
  
  # Create GOHyperGParams object
  params <- new("GOHyperGParams",
                geneIds = vnames,
                universeGeneIds = universe,
                annotation = "org.At.tair.db",
                ontology = "BP",
                pvalueCutoff = 0.01,
                conditional = FALSE,
                testDirection = "over")
  
  # Perform hypergeometric test
  hyper_test <- tryCatch({
    hyperGTest(params)
  }, error = function(e) {
    message("GO enrichment failed: ", e$message)
    return(NULL)
  })
  
  # If successful, format the results
  if (!is.null(hyper_test)) {
    go_table <- summary(hyper_test)
    go_table <- go_table[order(go_table$Pvalue), ]
    return(go_table[1:min(10, nrow(go_table)), ])
  } else {
    return(NULL)
  }
}
```

## Analyze Data Subsets

```{r analyze_function}
# Create a consolidated function for analyzing any subset
analyze_subset <- function(eset_data, subset_name) {
  # retreive expression data
  exp.norm <- exprs(eset_data)
  
  # Calculate mutual ranks
  mr_results <- calculate_mutual_ranks(exp.norm)
  gene.mr <- mr_results$mr
  
  # Get LFY networks at different MR thresholds
  cutoffs <- c(10, 20, 30, 50, 80, 100, 130, 210)
  lfy.mr.graphs <- lapply(cutoffs, get.mr.subgraph, mr.matrix = gene.mr, 
                          annotation = annotation, neighborhood = LFY)
  names(lfy.mr.graphs) <- sapply(lfy.mr.graphs, function(x) paste("lfy.mr.graph", x$cutoff, sep = "."))
  
  # Compare to known targets
  overlap.summary <- lapply(lfy.mr.graphs, evaluate_target_overlap, 
                           targets = targets, inf_targets = inf_targets, 
                           seedling_targets = seedling_targets)
  overlap.summary <- do.call(cbind, overlap.summary)
  colnames(overlap.summary) <- names(lfy.mr.graphs)
  
  # GO enrichment for each network
  go_results <- list()
  for (i in seq_along(lfy.mr.graphs)) {
    go_results[[names(lfy.mr.graphs)[i]]] <- perform_go_enrichment(lfy.mr.graphs[[i]], rownames(exp.norm))
  }
  
  # Display results
  cat(paste0("\n## Results for ", subset_name, "\n"))
  
  # Display overlap summary
  print(kable(t(overlap.summary), 
        caption = paste0("Overlap of MR networks with known LFY targets (", subset_name, ")")))
  
  # Display GO enrichment for MR = 80
  cat(paste0("\nGO Enrichment for MR = 80 (", subset_name, "):\n"))
  
  if (!is.null(go_results$lfy.mr.graph.80) && nrow(go_results$lfy.mr.graph.80) > 0) {
    print(kable(go_results$lfy.mr.graph.80, 
          caption = paste0("GO Enrichment for MR = 80 network (", subset_name, ")")))
  } else {
    cat("No significant GO enrichment found.\n")
  }
  
  # Return results
  return(list(graphs = lfy.mr.graphs, 
              overlap = overlap.summary, 
              go_enrichment = go_results,
              cutoffs = cutoffs))
}
```

## Apply Analysis to Each Subset

```{r analyze_subsets}
# Initialize results containers
subset_results <- list()

# Analyze Subset 1 if available
if (exists("eset_subset1")) {
  cat("\n### Analyzing Subset 1\n")
  subset_results$subset1 <- analyze_subset(eset_subset1, "Subset 1")
}

# Analyze Subset 2 if available
if (exists("eset_subset2")) {
  cat("\n### Analyzing Subset 2\n")
  subset_results$subset2 <- analyze_subset(eset_subset2, "Subset 2")
}

# Analyze Subset 3 if available
if (exists("eset_subset3")) {  
  cat("\n### Analyzing Subset 3\n")
  subset_results$subset3 <- analyze_subset(eset.small.2, "Subset 3")
}
```

# Visualize MR Networks

We'll visualize the network with the best enrichment for each subset.

```{r visualize_networks, fig.width=10, fig.height=8}
# Function to visualize network
visualize_mr_network <- function(lfy_graph, title) {
  plot(lfy_graph$graph, 
       layout = layout_with_fr(lfy_graph$graph),
       vertex.size = 8,
       vertex.label = V(lfy_graph$graph)$gene,
       vertex.label.cex = 0.7,
       edge.width = 1/E(lfy_graph$graph)$mr * 5,
       main = paste0(title, ", MR cutoff = ", lfy_graph$cutoff))
}

# Determine how many subsets we have
subset_names <- names(subset_results)
num_subsets <- length(subset_names)

if (num_subsets > 0) {
  # Arrange the plot layout based on number of subsets
  par(mfrow = c(1, min(3, num_subsets)))
  
  # Visualize each available network
  for (i in seq_along(subset_names)) {
    subset_name <- subset_names[i]
    display_name <- gsub("subset", "Subset ", subset_name)
    
    # Check if this subset has results and the MR=80 network
    if (length(subset_results[[subset_name]]) > 0 && 
        "lfy.mr.graph.80" %in% names(subset_results[[subset_name]]$graphs)) {
      visualize_mr_network(subset_results[[subset_name]]$graphs$lfy.mr.graph.80, display_name)
    }
  }
}
```

# Compare Results Across Subsets

```{r compare_results}
# Function to extract comparison metrics
extract_comparison_metrics <- function(results_list, cutoff = 80) {
  if (length(results_list) == 0) {
    return(NULL)
  }
  
  graph_name <- paste0("lfy.mr.graph.", cutoff)
  overlap <- results_list$overlap[, graph_name]
  
  go_enrichment <- results_list$go_enrichment[[graph_name]]
  if (!is.null(go_enrichment) && nrow(go_enrichment) > 0) {
    top_go <- go_enrichment$Term[1]
    top_go_pvalue <- go_enrichment$Pvalue[1]
  } else {
    top_go <- "None"
    top_go_pvalue <- NA
  }
  
  c(LFY_Nodes = overlap["LFY_nodes"],
    Percent_Known_Targets = overlap["PercentNetworkIsKnownAnyTarget"],
    Top_GO_Term = top_go,
    Top_GO_Pvalue = top_go_pvalue)
}

# Get list of subset names and create comparison table
subset_names <- names(subset_results)
display_names <- gsub("subset", "Subset ", subset_names)

# Initialize comparison table with the right dimensions
comparison_table <- data.frame(
  Subset = display_names,
  LFY_Nodes = NA,
  Percent_Known_Targets = NA,
  Top_GO_Term = NA,
  Top_GO_Pvalue = NA,
  stringsAsFactors = FALSE
)

# Fill in the comparison table
for (i in seq_along(subset_names)) {
  subset_name <- subset_names[i]
  
  # Extract metrics if results are available
  if (length(subset_results[[subset_name]]) > 0) {
    metrics <- extract_comparison_metrics(subset_results[[subset_name]])
    if (!is.null(metrics)) {
      comparison_table$LFY_Nodes[i] <- metrics["LFY_Nodes"]
      comparison_table$Percent_Known_Targets[i] <- metrics["Percent_Known_Targets"]
      comparison_table$Top_GO_Term[i] <- metrics["Top_GO_Term"]
      comparison_table$Top_GO_Pvalue[i] <- metrics["Top_GO_Pvalue"]
    }
  }
}

# Display comparison table
kable(comparison_table, 
      caption = "Comparison of MR = 80 Networks Across Subsets",
      row.names = FALSE)
```

# Save Results

```{r save_results}
# Save all results
all_results <- list(
  subset_results = subset_results,
  comparison = comparison_table
)

save(all_results, file = "../output/MR_ME00319_results.RData")

# Save individual subset results for easier access
for (subset_name in names(subset_results)) {
  if (length(subset_results[[subset_name]]) > 0) {
    subset_data <- subset_results[[subset_name]]
    save_file <- paste0("../output/MR_ME00319_", subset_name, ".RData")
    save(subset_data, file = save_file)
  }
}
```

# Conclusions

This analysis applies the Mutual Rank (MR) approach to the three subsets of the AtGenExpress developmental series dataset (ME00319) that were previously analyzed using WGCNA and correlation methods.

The results allow us to:

1. Compare the performance of MR across different data subsets
2. Determine which subset provides the most informative LFY network
3. Validate whether the MR approach consistently outperforms other methods regardless of the dataset used
4. Identify the optimal MR cutoff for each dataset

The comparative analysis with known targets and GO enrichment provides insights into which subset and cutoff are most effective for identifying biologically relevant LFY gene regulatory networks.
